---
title: "Ecosystem Service Climate Regulation in Cities | Norway"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cosmo
    code_folding: hide
---

```{css, echo=FALSE}
table {
  float: left;
}
```

Magnus Merkle (SSB)

Naturregnskapsprosjekt, 07. September 2026

\

# Changelog

Changes since version 1:

- Some text corrections
- Changed to 150m resolution as done in INCA
- Changed to Grunnkartet's CRS

# Abstract and Limitation

In this notebook we quantify the ecosystem service of local climate regulation (climate regulation in cities) for the Norwegian mainland in 2024. Most operations are executed with functions from R packages dplyr and tidyr (for general data formatting), terra (for geospatial data analysis), and stats (for regressions). The notebook is created on the basis of the following sources:

- INCA GIS Plugin Code Repository (2026). Last accessed 13 August 2026, the relevant script is here: <https://code.europa.eu/inca/gis-plugins/gis-plugins/-/blob/main/src/inca/localclimate/__init__.py>
- Marando et al. (2022) Urban heat island mitigation by green infrastructure in European Functional Urban Areas. Sustainable Cities and Society, 103564
- Eurostat (2026) Ecosystem accounts handbook, 2026 edition
- Immerzeel & Korkou (2024) Evaluation of the INCA-Tool for the purpose of ecosystem services accounting in Norway. EU Grants Project 2022-NO-EGD- Work Package 2. Norwegian Institute for Nature Research

The quantification process follows the "default approach" described in Eurostat's ecosystem accounts handbook. This approach requires less data capacity than the "advanced approach".

# 1. Filepaths and Functions

The main functions we use are part of R packages dplyr and tidyr, terra, and stats.

```{r, results = 'hide', message = FALSE, warning=FALSE}
# make space
rm(list = ls())
# packages
library(dplyr)
library(tidyr)
library(terra)
library(sf)
library(knitr)
library(viridisLite)
library(here)
library(stats)
library(tinytex)
library(osmdata)
# filepath
d_fp<-file.path("C:","Users","muk",
               paste0("OneDrive - Statistisk sentralbyr","\U00E5"),
               "data","INCA")
# file path to results
rfp<-file.path(here::here(),"results")
# A function to extract levels vectors from factor data stored in rasters 
# rst is raster file, lrn is layer name
gc()
```

# 2. Data Sources

Here is a table of data sources used in this project.

```{r}
dfs<-data.frame(matrix(nrow=7,ncol=8))
colnames(dfs) <- c("Input to","Data","Name","Provider","Resolution","Year","Future Updates","Notes")

dfs[1,]<-c("Aggregation",
           "Reference land cover map (LCM) with lookup table",
           "Grunnkart_OT",
           "Kartverket, NIBIO, SSB, Mdir",
           "100m grid",
           "2024",
           "2025",
           "...")
dfs[2,]<-c("Estimation, simulation, and aggregation",
           "City shapefiles",
           "DGURBA_RG_01M_2021_4258",
           "Eurostat (INCA default version 2026)",
           "Vector",
           "2021",
           "unknown",
           "The delineation of city boundaries is important because it defines how much of the surrounding greenspace is taken into account.")
dfs[3,]<-c("Estimation and simulation",
           "Local surface temperature data",
           "07_LST_2024_eu_masked",
           "Eurostat (INCA default version 2026)",
           "100m grid",
           "2024",
           "unknown",
           "This is based on data from USGS, processed by Eurostat to use for European ecosystem service accounting. The data has gaps for most of European cities. It represents (most likely) monthly median temperatures.")
dfs[4,]<-c("Estimation",
           "Tree cover density data.",
           "TCD_100m_2023_compressed",
           "Eurostat (INCA default version 2026)",
           "100m grid",
           "2023",
           "unknown",
           "This is based on data from the Chinese Academy of Sciences. We resample it to fit our 100m resolution grid.")
dfs[5,]<-c("Estimation",
           "Vegetation evaporation data.",
           "interc_transp_2024_Jun-Jul-Aug-Sept",
           "Eurostat (INCA default version 2026)",
           "150m grid",
           "2024",
           "unknown",
           "This is based on data from EU Space / Copernicus Land Monitoring Sevice")
dfs[6,]<-c("Estimation",
           "Weather station data",
           "StationNeighborhood2016",
           "Eurostat (INCA default version 2026)",
           "not gridded",
           "2016",
           "no",
           "This is weather station data from the US. Might be worth considering using weather data from Norway (due to different climate at the same latitudes as the US)")
dfs[7,]<-c("Aggregation",
           "NUTS2 shapefile for Norway",
           "NUTS2021_NO_LVL2_wos",
           "...",
           "Vector",
           "2021",
           "not relevant",
           "Only to retrieve results for specific subnational regions.")
#print
kable(dfs)
# clean up
rm(dfs)
```

# 3. Estimate a Model of LST for each City

## 3.1 Overview

In this first step we use gridded data to estimate a linear model of local surface temperature (LST) predicted by tree cover density (TCD) and evaporation. We estimate this model for each city, so it requires that our gridded data is cropped to city boundaries plus a buffer of 150 metres (as specified in Eurostat's handbook). The model is: $$ LST_{j} = \alpha_{1,j} + \beta_{1} TCD + \gamma_{1} EVAP $$ For each city, we save the estimated intercept of the linear model $\widehat{\alpha_{1,j}}$, which is equal to the LST in a scenario with no trees and evaporation. This is a purely hypothetical number with no interpretative value. It is needed as an intermediate step.

## 3.2 Import data

Read in city shapefiles, LST raster, TCD raster, EVAP raster. We also import Norwegian Land Cover data. Then crop all raster files to the Norwegian city boundaries with a 150m buffer.

We start by identifying cities in Norway that are counted within category "city" in the DGURBA dataset.

```{r,message = FALSE, results = 'hide'}
gc()
# read Grunnkartet -------------------------------------------------------------
lcm<-terra::rast(file.path(d_fp,"Norway_files","Grunnkart","Grunnkart_OT.tif"))

# read city shapefiles ---------------------------------------------------------
# Readme file in INCA's data repository states DEGURBA should be used for 2026
# reporting
cityb<-terra::vect(file.path(d_fp,"version_2026_7",
                             "inca_input_local_climate_regulation","Urban_Areas",
                             "2021","DEGURBA",
                             "DGURBA_RG_01M_2021_4258.shp"))
# MM: I checked the data and it has five categories. 1,2 and 3 relate to  degree 
# of urbanisation (1 is cities, 2 is towns and semi-dense areas, 3 is rural). 
# Additionally there is category 9 which exists only in Cyprus, so irrelevant 
# for us. And then there is category 0 which is used in lakes and such. Also not 
# relevant for us. So we subset the data to Norway only.
cityb<-cityb[cityb$CNTR_CODE == "NO",]
# And then we subset to feature category 1 urban only.
cityb<-cityb[cityb$DGURBA == 1,]
# For further operations we reproject the map to LCM's CRS and we add a 150m
# buffer to the LAU polygons, as suggested by the handbook
# INCA introduces the same buffer. See line 280 here:
# https://code.europa.eu/inca/gis-plugins/gis-plugins/-/blob/main/src/inca/localclimate/__init__.py?ref_type=heads
cityb <- terra::project(cityb,lcm)
cityb_buf <- terra::buffer(cityb,150)
# save
terra::writeVector(cityb,file.path(rfp,"no_city.shp"),overwrite=T)
terra::writeVector(cityb_buf,file.path(rfp,"no_citywbuffer.shp"),overwrite=T)
# print the city names
# unique(cityb$LAU_NAME)
```

As a result we get `r length(unique(cityb$LAU_NAME))` LAU in Norway that qualify as city. If we count Sandnes to Stavanger, and Bærum, Rælingen and Lørenskog to Oslo, we have four cities in total, namely Bergen, Oslo, Stavanger, and Trondheim. Let's have a look at the land cover map for each of the LAU to be considered. We translate the land cover map to the ecosystem typology required by Eurostat. Category 10 (marine inlets and transitional waters) is missing in the Norwegian LCM. This does not matter for the ecosystem service of local climate regulation as we only consider cooling from vegetation, according to the handbook.

```{r,message = FALSE, results = 'hide', fig.height = 9}
gc()
# then the recategorisation of grunnkartet
lcm_idf<-read.csv(file.path(d_fp,"Norway_files","Grunnkart",
                            "Lookup_grunnkart.csv"))
lcm_idf <- rbind(lcm_idf,c(NA,9900,"unclassified","uklassifisert"))
lcm_idf$landcover_id <- as.numeric(lcm_idf$landcover_id)
# make the vector of ecosystem types
estv<-c("Settlements and other artificial areas",
        "Cropland",
        "Grassland",
        "Forest and woodlands",
        "Heathlands and shrub",
        "Sparsely vegetated ecosystems",
        "Inland wetlands",
        "Rivers and canals",
        "Lakes and reservoirs",
        "Marine inlets and transitional waters",
        "Coastal beaches, dunes and wetlands",
        "Marine ecosystems")
# and add it to our lcm_idf
lcm_idf['es_type']<-NA
lcm_idf$es_type[lcm_idf$landcover_id %in% c(100,111,112,121,130,131,132,133,134,135,141,142,152)]<-estv[1] # settlem
lcm_idf$es_type[lcm_idf$landcover_id %in% c(200)]<-estv[2] # cropl
lcm_idf$es_type[lcm_idf$landcover_id %in% c(310,320)]<-estv[3] # grassl
lcm_idf$es_type[lcm_idf$landcover_id %in% c(400,410,420,440)]<-estv[4] # forest
lcm_idf$es_type[lcm_idf$landcover_id %in% c(520)]<-estv[5] # bush
lcm_idf$es_type[lcm_idf$landcover_id %in% c(600,611,623,631)]<-estv[6] # sparse
lcm_idf$es_type[lcm_idf$landcover_id %in% c(720)]<-estv[7] # wetlands
lcm_idf$es_type[lcm_idf$landcover_id %in% c(810)]<-estv[8] # rivers
lcm_idf$es_type[lcm_idf$landcover_id %in% c(910)]<-estv[9] # lakes
# category 10 (marine inlets and transitional waters) doesn't exist
lcm_idf$es_type[lcm_idf$landcover_id %in% c(1100)]<-estv[11] # coastal
lcm_idf$es_type[lcm_idf$landcover_id %in% c(1200)]<-estv[12] # marine ecosystems
#lcm_idf$es_type[lcm_idf$landcover_id %in% c(9900)]<-estv[12] # marine ecosystems
# make a numeric identifier also
lcm_idf$es_id <- match(lcm_idf$es_type, unique(estv))
# then add to raster
lev <- levels(lcm)[[1]]
lev$es_id <- lcm_idf$es_id[
  match(lev$Value, lcm_idf$landcover_id)
]
lev$es_type <- lcm_idf$es_type[
  match(lev$Value, lcm_idf$landcover_id)
]
levels(lcm) <- lev
# so here we can set which category we want to plot. 
# Can always see which categories we have available by calling cats(lcm)
activeCat(lcm) <- "es_type"
rm(lev)

# Make a plot of all cities
# import street data for better orientation
roads_list<-list()
for(i in 1:nrow(cityb_buf)){
  roads_list[[i]]<-terra::vect(file.path(here::here(),"ancillary_data",
                                         paste0("roads_",cityb_buf$LAU_ID[i],".shp")))
}
names(roads_list)<-cityb_buf$LAU_NAME
# Let's plot the data
for(i in 1:nrow(cityb_buf)){
  plot(mask(crop(lcm, cityb_buf[i,]),cityb_buf[i,]),main=paste(cityb_buf$LAU_NAME[i],"LCM"), legend=T)
  plot(roads_list[[i]],add=T)
  
}
```

Now next we need to identify which of these cities have experienced maximum daily temperatures of at least 25 degrees celsius in the reference year of 2024. For this purpose we first download and import Norwegian meteorological data for 2024 covering the four cities. For now I have eyeballed which weathers stations fit in which city. Given that we observe maximum temperatures of at least 25 degrees celsius in all of them, it does not really matter, the precise selection of weather stations does not really matter. Below we print the weather stations and months in which there where temperatures above 25 deg C.

```{r,message = FALSE, results = 'hide'}
gc()
# read in meteorogogical data for stations inside the above city LAUs for 2024 -
tempd <- read.csv(file.path(here::here(),"ancillary_data",
                            "NO_2024_citylau_dailymaxtemp.csv"),sep=";")
colnames(tempd)<-c("station_name","station_code","date","maxtemp")
tempd$maxtemp <- sub(",",".",tempd$maxtemp )
tempd$maxtemp <- as.numeric(tempd$maxtemp)
# save all cities with daily temperatures above 25 degrees
stations<-tempd %>% filter(maxtemp >= 25) %>% select(station_name) %>% unique()
# check for months in each city
heatmonths<-list()
for (i in 1:nrow(stations)){
  heatmonths[i]<- tempd %>% filter(maxtemp >= 25) %>% filter(station_name==stations[i,1]) %>% select(date) %>% unique()
}
names(heatmonths)<-stations[,1]
for(i in 1:nrow(stations)){
  print(stations[i,1])
  print(unique(substr(heatmonths[[i]],4,5)))
}
# clean up
rm(stations,tempd,heatmonths)
```

So this shows that June and July are the months where we have observed daily max temps exceeding 25 degrees in all weather stations within the cities in question in 2024. August and September are months in which some city LAU have observed max temps exceeding 25 degrees. The Eurostat handbook suggests to use an inclusive approach. We therefore pick all available months for the LST data, which we read in next. Croping and reprojection is done outside of this script, because it is a computationally time-intensive process

```{r,message = FALSE, results = 'hide'}
gc()
# read LST raster file ---------------------------------------------------------
# Here we have data available for June, July, August, September, and we pick
# as described above.
lstr_6_no<-terra::rast(file.path(here::here(),"ancillary_data",
                                 "lstr_6_no.tif"))
lstr_7_no<-terra::rast(file.path(here::here(),"ancillary_data",
                                 "lstr_7_no.tif"))
lstr_8_no<-terra::rast(file.path(here::here(),"ancillary_data",
                                 "lstr_8_no.tif"))
lstr_9_no<-terra::rast(file.path(here::here(),"ancillary_data",
                                 "lstr_9_no.tif"))

# mask to city LAUs
lstr_6_no <- mask(crop(lstr_6_no,cityb_buf),cityb_buf)
lstr_7_no <- mask(crop(lstr_7_no,cityb_buf),cityb_buf)
lstr_8_no <- mask(crop(lstr_8_no,cityb_buf),cityb_buf)
lstr_9_no <- mask(crop(lstr_9_no,cityb_buf),cityb_buf)

## MM comment: The unit of the data is Kelvin. Vito has applied the scaling factor
# that is mentioned here: 
# https://www.usgs.gov/landsat-missions/landsat-collection-2-surface-temperature
# You can trace the step in line 131 here
# https://code.europa.eu/inca/gis-plugins/gis-plugins/-/blob/main/src/inca/localclimate/preprocessing/preprocess_LST.js?ref_type=heads
# Before running regressions, VITO changes the unit to celsuis. See line 195 here:
# https://code.europa.eu/inca/gis-plugins/gis-plugins/-/blob/main/src/inca/localclimate/__init__.py?ref_type=heads
# so we do that here too.
lstr_6_no <- lstr_6_no - 273.15
lstr_7_no <- lstr_7_no - 273.15
lstr_8_no <- lstr_8_no - 273.15
lstr_9_no <- lstr_9_no - 273.15

# Move LST data all into one stack
lstr_stack <- c(lstr_6_no,lstr_7_no,lstr_8_no,lstr_9_no)
rm(lstr_6_no,lstr_7_no,lstr_8_no,lstr_9_no)
```

Then plot the LST data, to show that it has gaps. We use the land cover map as background. First for June:

```{r,message = FALSE, results = 'hide', fig.height = 9}
gc()
# Let's plot the data
for(i in 1:nrow(cityb_buf)){
plot(mask(crop(lcm, cityb_buf[i,]),cityb_buf[i,]),main=paste(cityb_buf$LAU_NAME[i],"June"), legend=F)
plot(mask(crop(lstr_stack[[1]], cityb_buf[i,]),cityb_buf[i,]),add=T)
plot(roads_list[[i]],add=T)
}
```

You see the largets data gaps for Sandnes and Bergen. Also some gaps for Trondheim and Stavanger. Then look at the maps for July:

```{r,message = FALSE, results = 'hide', fig.height = 9}
gc()
# Let's plot the data
for(i in 1:nrow(cityb_buf)){
plot(mask(crop(lcm, cityb_buf[i,]),cityb_buf[i,]),main=paste(cityb_buf$LAU_NAME[i],"July"), legend=F)
plot(mask(crop(lstr_stack[[2]], cityb_buf[i,]),cityb_buf[i,]),add=T)
plot(roads_list[[i]],add=T)
}
```

So for month July the largest gaps prevail in the LST data for Sandnes, Trondheim, and Bærum. Also substantial gaps for Oslo and Bergen, and some for Stavanger, Rælingen, and Lørenskog. The gaps do not look much different in the data for the other available months, so I omit plotting those here. I was wondering whether they could occur due to clouds, blanking out parts of the original satellite picture? On the other hand, I notice that some of the gaps look quite similar across months, which makes the cloud hypothesis less likely. In any case, what we could do in a next version is to compute median or mean values across months first, so we get an LST map with the best possible coverage. Given that INCA only reads in one single LST file at a time, I currently stick the seperate month-wise calculations, to allow comparison of resuts to INCA outputs.

Then we read in the other raster files needed. Tree cover density and then evaporation. For evaporation we have the choice between different datasets. One covering July to August, one covering June to August, and a third covering June to September. There is no guidance in the Eurostat handbook on which data to choose. To match the LST data selection, we choose to use June to September.

```{r, message = FALSE, results = 'hide'}
gc()
# rough crop
norcrop <- ext(
  3.5e6, 5.5e6, # xmin, xmax
  3.5e6, 5.5e6 # ymin, ymax
)

# read EVAP raster file --------------------------------------------------------
evapr<-terra::rast(file.path(d_fp,"version_2026_7",
                           "inca_input_local_climate_regulation",
                           "Vegetation_Evapotranspiration","2024",
                           "interc_transp_2024_Jun-Jul-Aug-Sept.tif"))
# I have found one funny thing here. The metadata in this raster file includes a
# scale factor of 0.01. R terra automatically applies this to the values of the 
# file, so it contains realistic numbers. But the python code in INCA does not
# apply the scale factor. Therefore, numbers in INCA for evapotranspiration are 
# 100 times larger than here. For now, I manually change this here, so we have
# results as close.

# rough crop
evapr_no <- crop(evapr,norcrop)
# reproject
evapr_no <- terra::project(evapr_no,crs(lcm),res=res(evapr_no),method="bilinear")
# mask to city lau
evapr_no <- mask(crop(evapr_no,cityb_buf),cityb_buf)
evapr_no <- 100*evapr_no # this step to immitate what INCA does (but it actually
# is not correct)
# When plotting the resulting map we see that it is a bit patchy, too. Possibly
# revisit later.


# read TCD raster file ---------------------------------------------------------
tcdr<-terra::rast(file.path(d_fp,"version_2026_7",
                           "inca_input_local_climate_regulation",
                           "Tree_Cover_Density","2023",
                           "TCD_100m_2023_compressed.tif"))
# rough crop
tcdr_no <- crop(tcdr,norcrop)
# reproject
tcdr_no <- terra::project(tcdr_no,evapr_no)
# mask to city lau
tcdr_no <- mask(crop(tcdr_no,cityb_buf),cityb_buf)
# MM I have checked the resulting data and it looks fine. No gaps.

# reproject lst raster data ----------------------------------------------------
lstr_stack<-terra::project(lstr_stack,evapr_no)

# tidy up
rm(tcdr,evapr)
```

## 3.3 Run regressions

We loop across cities and across available months here. In each city and month we run a regression where grid cells are units of observation. Predicting LST with TCD and EVAP. We store the resulting intercepts for each city and month.

```{r, message = FALSE, results = 'hide'}
gc()
# months to consider
mdr<-c(6,7,8,9)

# our dataframe for results
reg1df<-as.data.frame(matrix(nrow=1,ncol=14))
colnames(reg1df)<-c("month","city","pixels_in_shp","p_pixels_used","p_pixels_rm",
                    "p_pixels_rm_due_na_lcm","p_pixels_rm_due_na_data","avg_lst",
                    "avg_tcd","avg_evap","est_intercept","est_tcd","est_evap",
                    "R2")
# function
first_regression_function<-function(cityshp,lst_rfile,tcd_rfile,evap_rfile){
  # Step 1: Crop the data to the right shape, stack it, and turn into df
  estack<-c(mask(crop(lst_rfile, cityshp),cityshp),
            mask(crop(tcd_rfile, cityshp),cityshp),
            mask(crop(evap_rfile, cityshp),cityshp))
  names(estack)<-c("lst","tcd","evap")
  edf<-as.data.frame(estack)
  clean_edf <- edf %>% drop_na()
  rm(estack)
  # Step 2: Run regression
  elm <- lm(lst ~ tcd + evap, data = edf)
  # Step 3: Save relevant info
  rv<-vector()
  # total pixel number in shape
  rv[1] <- mask(crop(init(lst_rfile, 1), cityshp), cityshp) %>% ncell()
  # proportion of pixels used
  rv[2] <- nobs(elm) / rv[1]
  # proportion of pixels removed
  rv[3] <- 1 - rv[2]
  # proportion of pixels removed because outside of lcm data
  rv[4] <- 1 - (nrow(edf) / rv[1])
  # proportion of pixels removed because of data gaps (lst,tcd,evap)
  rv[5] <- (nrow(edf) - nobs(elm)) / rv[1]
  # average land surface temp
  rv[6] <- mean(clean_edf$lst)
  # average tree cover density
  rv[7] <- mean(clean_edf$tcd)
  # average evapotranspiration
  rv[8] <- mean(clean_edf$evap)
  # estimated intercept
  rv[9] <- elm$coefficients[1]
  # estimated coefficient on tre cover density
  rv[10] <- elm$coefficients[2]
  # estimated coefficient on evaporation
  rv[11] <- elm$coefficients[3]
  # model fit
  rv[12] <- summary(elm)$r.squared
  # clean up and return result
  rm(edf,clean_edf,elm)
  return(rv)
}

## then loop across months and cities
rl<-list()
for(j in 1:length(mdr)){
  rl[[j]]<-reg1df
  for(i in 1:nrow(cityb_buf)){
    rl[[j]][i,1]<-mdr[j]
    rl[[j]][i,2]<-as.data.frame(cityb_buf[i,])$LAU_NAME
    rl[[j]][i,3:14]<-first_regression_function(cityshp = cityb_buf[i,],
                                lst_rfile = lstr_stack[[j]],
                                tcd_rfile = tcdr_no,
                                evap_rfile = evapr_no)
  }
}
firstdf<-do.call("rbind",rl)
# add city codes
firstdf <- firstdf %>% 
  mutate(city_code = cityb_buf$LAU_ID[match(city,cityb_buf$LAU_NAME)]) %>%
  relocate(city_code,.after=city)
#clean up
rm(reg1df,rl)

# save coefficients in raster stacks
# one raster per month
intercept_rasters <- vector("list", length(mdr))
beta1_rasters <- vector("list", length(mdr))
gamma1_rasters <- vector("list", length(mdr))
# loop across months
for (j in seq_along(mdr)) {
    # subset results for current month
  month_df <- subset(firstdf, month == mdr[j])
    # copy polygons
  city_month <- cityb_buf
    # attach intercepts and convert to degree celsius
  city_month$est_intercept <- month_df$est_intercept[
    match(city_month$LAU_NAME, month_df$city)
    ]
  city_month$est_tcd <- month_df$est_tcd[
    match(city_month$LAU_NAME, month_df$city)
    ]
  city_month$est_evap <- month_df$est_evap[
    match(city_month$LAU_NAME, month_df$city)
    ]
    # rasterize
  intercept_rasters[[j]] <- terra::rasterize(
    city_month,
    lstr_stack[[1]], # as a template
    field = "est_intercept"
  )
  beta1_rasters[[j]] <- terra::rasterize(
    city_month,
    lstr_stack[[1]], # as a template
    field = "est_tcd"
  )
  gamma1_rasters[[j]] <- terra::rasterize(
    city_month,
    lstr_stack[[1]], # as a template
    field = "est_evap"
  )
  rm(city_month,month_df)
}
# create SpatRaster stack
intercept_stack <- rast(intercept_rasters)
beta1_stack <- rast(beta1_rasters)
gamma1_stack <- rast(gamma1_rasters)
names(intercept_stack) <- paste0("intercept_", mdr)
names(beta1_stack) <- paste0("est_tcd_", mdr)
names(gamma1_stack) <- paste0("est_evap_", mdr)
# tidy up
rm(intercept_rasters,beta1_rasters,gamma1_rasters)
```

Print the table, with rounded values.

```{r}
# print result table
kable(firstdf %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```

# 4. Estimate a Model of Air Temperature

## 4.1 Overview

In this second step we use weather station data to estimate a linear model of maximum air temperature predicted by local surface temperature and latitude. Here is the unit of observation is weather stations. We save the estimated coefficients.

## 4.2 Import data

Read in a single dataframe containing weather station data. This is the training data from the US.

```{r, message = FALSE, results = 'hide'}
gc()
# read in weather station data
wstd<-read.csv(file.path(dfp,"version_2026_7",
                           "inca_input_local_climate_regulation",
                           "Weather_station_data",
                           "StationNeighborhood2016.csv"))
```

## 4.3 Run regression

A single regression to estimate a linear model. Predicting maximum air temperature with local surface temperature and latitude. The model we estimate is: $$ T^{air} = \beta_{2} LST + \gamma_{2}LAT $$

```{r, message = FALSE, results = 'hide'}
gc()
# So here we now estimate the model and store the coefficients and adjusted r squared
wlm <- lm(TempMax ~ st + Ycoor, data = wstd)
beta2h <- wlm$coefficients['st']
gamma2h <- wlm$coefficients['Ycoor']
wlm_adjr2<-summary(wlm)$adj.r.squared
# tidy up
rm(wstd)
```

Here are model outputs:

```{r}
summary(wlm)
rm(wlm)
```

# 5. Simulate Air Temperature for the City Grids

## 5.1 Overview

Here we now apply the estimated coefficients from the previous two steps to generate simulated gridded data. We do this for each city. We first need to construct predicted land surface temperature with our calibrated model and observed inputs for the right hand side.

$$ \widehat{LST}^{med.veg}_{j,i} = \widehat{\alpha}_{1,j} + \widehat{\beta}_{1,j}TCD_{j,i} + \widehat{\gamma}_{1,j}EVAP_{ji}$$

Then we compute for each grid cell $i$ the expected air temperature given observed conditions.

$$ \widehat{T^{air}_{j,i}} = \widehat{\beta_{2}} \widehat{LST}^{med.veg}_{j,i} + \widehat{\gamma_{2}} latitude_{j,i} $$.

Then we compute for each grid cell $i$ the expected air temperature in a no-vegetation scenario, i.e. where $\widehat{LST^{uten.veg}_{j,i}} = \widehat{\alpha_{1,j}}$ because $TCD = EVAP = 0$ in this case. So

$$ \widehat{T^{air,0}_{ji}} = \widehat{\beta_{2}} \widehat{LST^{uten.veg}_{j,i}} + \widehat{\gamma_{2}} latitude_{j,i} $$.

The cooling effect is then retrieved for each grid cell $i$ in city $j$ as

$$ \widehat{cooling_{j,i}} = \widehat{T^{air,0}_{ji}} - \widehat{T^{air}_{ji}}$$.

After plugging in for estimated grid-cell temperature in both secenarios with vegetation and without vegetation, the previous equation simplifies to:

$$ \widehat{cooling}_{j,i} = -\widehat{\beta}_{2} * (\widehat{\beta}_{1,j}TCD_{j,i} + \widehat{\gamma}_{1,j}EVAP_{ji})$$

## 5.2 Create cooling rasters

Make a loop where we compute the cooling rasters for each city in each month. Based on the method, it is possible to get positive and negative cooling values for pixels. The guidance says nothing about that, but the python script behind INCA reveals that negative cooling values are forced to zero (line 512 in the python script "__init___.py"). So what we do is to create two versions, one with negative values removed, one that keeps negative values. And then we create raster maps of cooling. 
```{r,message = FALSE, results = 'hide'}
gc()
# first we create a latitude raster
# Project raster to geographic coordinates
tmp_ll <- project(lstr_stack[[1]], "EPSG:4326")
# Create latitude raster
lat_ll <- init(tmp_ll, "y")
# Back to the original grid
lat_r <- project(lat_ll, lstr_stack[[1]])
rm(tmp_ll,lat_ll)
# then compute the cooling rasters in a loop across months
# the loop
cooling_rasters <- vector("list", length(mdr))
cooling_rasters_nn <- vector("list", length(mdr)) # non-negative version as in INCA
# loop across months
for (j in seq_along(mdr)) {
  # modelled cooling
  coolingr <- (-beta2h) * ((beta1_stack[[j]]*tcdr_no) + (gamma1_stack[[j]]*evapr_no))
  coolingr_nn <- coolingr
  coolingr_nn[coolingr_nn < 0] <- 0
  # save
  cooling_rasters[[j]] <- coolingr
  cooling_rasters_nn[[j]] <- coolingr_nn
  writeRaster(coolingr,file.path(rfp,paste0("cooling_",mdr[j],".tif")),overwrite=T)
  writeRaster(coolingr_nn,file.path(rfp,paste0("cooling_nn_",mdr[j],".tif")),overwrite=T)
  # tidy up
  rm(coolingr,coolingr_nn)
}
# create SpatRaster stack
cooling_stack <- rast(cooling_rasters)
cooling_nn_stack <- rast(cooling_rasters_nn)
names(cooling_stack) <- paste0("Cooling_", mdr)
names(cooling_nn_stack) <- paste0("Cooling_", mdr)

# tidy up
rm(lat_r,cooling_rasters)
```


And then plot results. Here for June:
```{r,message = FALSE, results = 'hide', fig.height = 9}
# now for plots and aggregations need to align extents perfectly
lcm <- resample(
  lcm,
  cooling_stack[[j]],
  method = "near"
)
# need to recode lcm raster
rcl <- as.matrix(cats(lcm)[[1]][, c("Value", "es_id")])
lcm2 <- classify(lcm, rcl)
# then plot for June
for(j in 1){
  for(i in 1:nrow(cityb_buf)){
    plot(mask(crop(lcm, cityb_buf[i,]),cityb_buf[i,]),main=paste(cityb_buf$LAU_NAME[i],"month",mdr[j]), legend=F)
    plot(mask(crop(cooling_stack[[j]], cityb_buf[i,]),cityb_buf[i,]),add=T)
    plot(roads_list[[i]],add=T)
  }
}
```

And then plot July

```{r,message = FALSE, results = 'hide', fig.height = 9}
# then plot for July
for(j in 2){
  for(i in 1:nrow(cityb_buf)){
    plot(mask(crop(lcm, cityb_buf[i,]),cityb_buf[i,]),main=paste(cityb_buf$LAU_NAME[i],"month",mdr[j]), legend=F)
    plot(mask(crop(cooling_stack[[j]], cityb_buf[i,]),cityb_buf[i,]),add=T)
    plot(roads_list[[i]],add=T)
  }
}
```

# 6. Summarise results

Here we now summarise the data across grid cells (and time). The final outputs are SUT tables. We compute average cooling by ecosystem type and area, and weighted average cooling across all ecosystem types by area. INCA reports the city-level results based on non-modified cooling data (where negative cooling isn't forced to zero) while the NUTS-level results are reported based on modified cooling data (negative cooling means zero cooling). So we do the same here.

## 6.1 Supply tables for City LAUs in Norway

```{r,message = FALSE, results = 'hide'}
gc()
### Make a function and then run for each month
cooling_by_city<-function(j){
  df_sut<-as.data.frame(matrix(ncol=14,nrow=nrow(cityb_buf)))
  colnames(df_sut)<-c("LAU",estv,"w_avg_cool")
  # now loop across LAU
  for(i in 1:nrow(cityb_buf)){
  # the LAU name
  df_sut$LAU[i]<-as.data.frame(cityb_buf[i,])$LAU_NAME
  # compute cooling averages
  # using terra zonal
  z <- terra::zonal(mask(crop(cooling_stack[[j]], cityb_buf[i,]),cityb_buf[i,]), # values
                    mask(crop(lcm2, cityb_buf[i,]),cityb_buf[i,]),# categories
                    fun="mean",
                    na.rm=T)
  # create city-masked cooling raster
  cool <- mask(crop(cooling_stack[[j]], cityb_buf[i,]), cityb_buf[i,])
  # mask the lcm for the city cooling (to exclude cells that are NA in cooling)
  lcm_city <- mask(crop(lcm2, cityb_buf[i,]), cityb_buf[i,])
  cool_lcm <- mask(lcm_city, cool)
  # set column names
  colnames(z) <- c("es_id","cooling")
  # add ecosystem names
  z$es_name <- lcm_idf$es_type[
  match(z$es_id, lcm_idf$es_id)]
  # here is the weighted cooling across all ecosystems
  dfp<-freq(cool_lcm)
  dfp['es_weight']<-dfp$count / dfp %>% select(count) %>% sum()
  z <- z %>% left_join(dfp,by=join_by("es_id" == "value"))
  z['weighted_average_cooling']<-z$cooling*z$es_weight
  rm(dfp,cool,lcm_city,cool_lcm)
  # paste result into our table, loop again, just to ensure we keep empty values where ecosystems do not exist
  for(k in 1:length(estv)){
    df_sut[i,k+1]<- {
      x <- z %>% filter(es_name==estv[k]) %>% pull(cooling)
      if(length(x) == 0) NA else x
    }
    rm(x)
  }
  # finally the weighted average for the LAU as a whole
  df_sut$w_avg_cool[i]<-sum(z$weighted_average_cooling)
  rm(z)
  }
#return result
return(df_sut)
}

# then run
city_st_6 <- cooling_by_city(j=1)
city_st_7 <- cooling_by_city(j=2)
city_st_8 <- cooling_by_city(j=3)
city_st_9 <- cooling_by_city(j=4)
```

Here are the results for June (rounded)

```{r}
kable(city_st_6 %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```

Here are the results for July (rounded to two digits)

```{r}
kable(city_st_7 %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```

## 6.2 Supply tables by NUTS2 and 0 in Norway

```{r,message = FALSE, results = 'hide'}
gc()
# Import Norway NUTs2 shape
n2shp<-terra::vect(file.path(d_fp,"Norway_files","NUTS2021_NO_LVL2_wosvalbard","NUTS2021_NO_LVL2_wos.shp"))
n2shp <- project(n2shp,crs(lcm))

# make a function
cooling_by_nuts<-function(j){
  df_sut<-as.data.frame(matrix(ncol=14,nrow=nrow(n2shp)))
  colnames(df_sut)<-c("Region",estv,"w_avg_cool")
  # now loop across NUTS2
  for (i in 1:nrow(n2shp)){
  # the NUTS name
  df_sut$Region[i]<-as.data.frame(n2shp[i,])$NUTS_ID
  # compute cooling averages
  # using terra zonal
  if(df_sut$Region[i]!="NO07"){ # if not nord norge (for which we do not have any data)
    z <- terra::zonal(mask(crop(cooling_nn_stack[[j]], n2shp[i,]),n2shp[i,]), # values
                      mask(crop(lcm2, n2shp[i,]),n2shp[i,]),# categories
                      fun="mean",
                      na.rm=T)
    # create a mask for cities within the shape
    n2_city_buf<-mask(crop(cityb_buf,n2shp[i,]),n2shp[i,])
    #create region-masked cooling raster
    cool <- mask(crop(cooling_nn_stack[[j]], n2_city_buf), n2_city_buf)
    # mask the lcm for the city cooling (to exclude cells that are NA in cooling)
    lcm_nuts <- mask(crop(lcm2, n2_city_buf), n2_city_buf)
    cool_lcm <- mask(lcm_nuts, cool)
    # set column names
    colnames(z) <- c("es_id","cooling")
    # add ecosystem names
    z$es_name <- lcm_idf$es_type[
      match(z$es_id, lcm_idf$es_id)]
    # if there is any cooling in the shape
    if(any(!is.na(z$cooling))){
      dfp<-freq(cool_lcm)
      dfp['es_weight']<-dfp$count / dfp %>% select(count) %>% sum()
      z <- z %>% left_join(dfp,by=join_by("es_id" == "value"))
      z['weighted_average_cooling']<-z$cooling*z$es_weight
      rm(dfp)
      # paste result into our table, loop again, just to ensure we keep empty values where ecosystems do not exist
      for(k in 1:length(estv)){
        df_sut[i,k+1]<- {
          x <- z %>% filter(es_name==estv[k]) %>% pull(cooling)
          if(length(x) == 0) NA else x
        }
        rm(x)
      }
      # finally the weighted average for the LAU as a whole
      df_sut$w_avg_cool[i]<-sum(z$weighted_average_cooling,na.rm=T)
      rm(z,n2_city_buf,cool,cool_lcm)
    }
  }
  }
  # now the process for all of norway (MM TO REVISE SIMILAR AS ABOVE)
  df_sut<-rbind(df_sut,c("NO",rep(NA,13)))
  z <- terra::zonal(mask(crop(cooling_nn_stack[[j]], n2shp),n2shp), # values
                  mask(crop(lcm2, n2shp),n2shp),# categories
                  fun="mean",
                  na.rm=T)
  # set column names
  colnames(z) <- c("es_id","cooling")
  # add ecosystem names
  z$es_name <- lcm_idf$es_type[
  match(z$es_id, lcm_idf$es_id)]
  # if there is any cooling in the shape
  if(any(!is.na(z$cooling))){
  # we first need to cut out the right city LAUs
  dfp<-freq(mask(crop(lcm2, cityb_buf),cityb_buf))
  dfp['es_weight']<-dfp$count / freq(mask(crop(lcm2, cityb_buf),cityb_buf)) %>% select(count) %>% sum()
  z <- z %>% left_join(dfp,by=join_by("es_id" == "value"))
  z['weighted_average_cooling']<-z$cooling*z$es_weight
  rm(dfp)
  # paste result into our table, loop again, just to ensure we keep empty values where ecosystems do not exist
  for(k in 1:length(estv)){
    df_sut[which(df_sut$Region=="NO"),k+1]<- {
      x <- z %>% filter(es_name==estv[k]) %>% pull(cooling)
      if(length(x) == 0) NA else x
    }
    rm(x)
  }
  # finally the weighted average for the LAU as a whole
  df_sut$w_avg_cool[df_sut$Region=="NO"]<-sum(z$weighted_average_cooling,na.rm=T)
  rm(z)
  }
  # turn all columns but the first into numeric and return results
  df_sut <- df_sut %>% mutate(across(-1, as.numeric))
  return(df_sut)
}

# then run
nuts_st_6 <- cooling_by_nuts(j=1)
nuts_st_7 <- cooling_by_nuts(j=2)
nuts_st_8 <- cooling_by_nuts(j=3)
nuts_st_9 <- cooling_by_nuts(j=4)
```

Here are the results for June (rounded)

```{r}
kable(nuts_st_6 %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```

Here are the results for July (rounded)

```{r}
kable(nuts_st_7 %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```

## 6.3 Use tables

```{r}
ut_6 <- nuts_st_6 %>% 
  select(c(Region,w_avg_cool)) %>% 
  mutate(`Intermediate consumption by industries` = NA) %>%
  mutate(`Government final consumption` = NA) %>%
  mutate(`Households final consumption` = w_avg_cool) %>%
  mutate(`Gross capital formation` = NA) %>%
  mutate(Exports = NA)
ut_7 <- nuts_st_7 %>% 
  select(c(Region,w_avg_cool)) %>% 
  mutate(`Intermediate consumption by industries` = NA) %>%
  mutate(`Government final consumption` = NA) %>%
  mutate(`Households final consumption` = w_avg_cool) %>%
  mutate(`Gross capital formation` = NA) %>%
  mutate(Exports = NA)
ut_8 <- nuts_st_8 %>% 
  select(c(Region,w_avg_cool)) %>% 
  mutate(`Intermediate consumption by industries` = NA) %>%
  mutate(`Government final consumption` = NA) %>%
  mutate(`Households final consumption` = w_avg_cool) %>%
  mutate(`Gross capital formation` = NA) %>%
  mutate(Exports = NA)
ut_9 <- nuts_st_9 %>% 
  select(c(Region,w_avg_cool)) %>% 
  mutate(`Intermediate consumption by industries` = NA) %>%
  mutate(`Government final consumption` = NA) %>%
  mutate(`Households final consumption` = w_avg_cool) %>%
  mutate(`Gross capital formation` = NA) %>%
  mutate(Exports = NA)
```

Here are the results for June (rounded)

```{r}
kable(ut_6 %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```

Here are the results for July (rounded)

```{r}
kable(ut_6 %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```
