---
title: "Ecosystem Service Climate Regulation in Cities | Norway"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cosmo
    code_folding: hide
---

```{css, echo=FALSE}
table {
  float: left;
}
```

Magnus Merkle (SSB)

Naturregnskapsprosjekt, 1. October 2026

Please do not share. This is work in progress.

\

# Changelog

Changes since version 3:

- Removed truncation of negative cooling
- Changed LST inputs to a seasonally adjusted median LST across relevant months
- Removed explicit INCA comparisons, to simplify the document
- Aligned the city-wise calculation to the region-wise calculation for consistency (it is inconsistent in INCA)
- Reduced the number of printed results to only show what we report


# Abstract and Limitation

In this notebook we quantify the ecosystem service of local climate regulation (climate regulation in cities) for the Norwegian mainland in 2024. Most operations are executed with functions from R packages dplyr and tidyr (for general data formatting), terra (for geospatial data analysis), and stats (for regressions). The notebook is created on the basis of the following sources:

- INCA GIS Plugin Code Repository (2026). Last accessed 13 August 2026, the relevant script is here: <https://code.europa.eu/inca/gis-plugins/gis-plugins/-/blob/main/src/inca/localclimate/__init__.py>
- Marando et al. (2022) Urban heat island mitigation by green infrastructure in European Functional Urban Areas. Sustainable Cities and Society, 103564
- Eurostat (2026) Ecosystem accounts handbook, 2026 edition
- Immerzeel & Korkou (2024) Evaluation of the INCA-Tool for the purpose of ecosystem services accounting in Norway. EU Grants Project 2022-NO-EGD- Work Package 2. Norwegian Institute for Nature Research
- Hellman, F., Brown, N. & Moström, J. (2025) Ecosystem accounts in Sweden. Deliverable 2.1 of Eurostat grant project 2023-SE-EGD: Advanced data modelling for ecosystem accounts. Statistics Sweden (SCB), Environmental Accounts.
- Bellingen, M. et al.(2026) Ecosystem Service Account for Germany. Deliverable 2.2 Methodological report for the ecosystem service local climate regulation. Federal Statistical Office (DESTATIS).

The quantification process follows the "default approach" described in Eurostat's ecosystem accounts handbook. This approach requires less data capacity than the "advanced approach".

# 1. Filepaths and Functions

The main functions we use are part of R packages dplyr and tidyr, terra, and stats.
```{r, results = 'hide', message = FALSE, warning=FALSE}
# make space
rm(list = ls())
# packages
library(dplyr)
library(tidyr)
library(terra)
library(sf)
library(knitr)
library(viridisLite)
library(here)
library(stats)
library(tinytex)
library(osmdata)
# filepath
d_fp<-file.path("C:","Users","muk",
               paste0("OneDrive - Statistisk sentralbyr","\U00E5"),
               "data","INCA")
# file path to results
rfp<-file.path(here::here(),"results")
# A function to extract levels vectors from factor data stored in rasters 
# rst is raster file, lrn is layer name
gc()
```

# 2. Data Sources

Here is a table of data sources used in this project.
```{r}
dfs<-data.frame(matrix(nrow=10,ncol=8))
colnames(dfs) <- c("Input to","Data","Name","Provider","Resolution","Year","Future Updates","Notes")

dfs[1,]<-c("Aggregation",
           "Reference land cover map (LCM) with lookup table",
           "Grunnkart_OT",
           "Kartverket, NIBIO, SSB, Mdir",
           "100m grid",
           "2024",
           "2025",
           "...")
dfs[2,]<-c("Estimation, simulation, and aggregation",
           "City shapefiles",
           "DGURBA_RG_01M_2021_4258",
           "Eurostat (INCA default version 2026)",
           "Vector",
           "2021",
           "unknown",
           "The delineation of city boundaries is important because it defines how much of the surrounding greenspace is taken into account.")
dfs[3,]<-c("Estimation and simulation",
           "Local surface temperature data",
           "MM_LST_2024_eu_masked",
           "Eurostat (INCA default version 2026)",
           "100m grid",
           "2024",
           "unknown",
           "This is based on data from USGS, processed by Eurostat to use for European ecosystem service accounting. The data has gaps for most of European cities. It represents (most likely) monthly median temperatures.")
dfs[4,]<-c("Estimation",
           "Tree cover density data.",
           "TCD_100m_2023_compressed",
           "Eurostat (INCA default version 2026)",
           "100m grid",
           "2023",
           "unknown",
           "This is based on data from the Chinese Academy of Sciences. We resample it to fit our 100m resolution grid.")
dfs[5,]<-c("Estimation",
           "Vegetation evaporation data.",
           "interc_transp_2024_Jun-Jul-Aug-Sept",
           "Eurostat (INCA default version 2026)",
           "150m grid",
           "2024",
           "unknown",
           "This is based on data from EU Space / Copernicus Land Monitoring Sevice")
dfs[6,]<-c("Estimation",
           "Weather station data",
           "StationNeighborhood2016",
           "Eurostat (INCA default version 2026)",
           "not gridded",
           "2016",
           "no",
           "This is weather station data from the US. Might be worth considering using weather data from Norway (due to different climate at the same latitudes as the US)")
dfs[7,]<-c("Aggregation",
           "NUTS2 shapefile for Norway",
           "NUTS2021_NO_LVL2_wos",
           "...",
           "Vector",
           "2021",
           "not relevant",
           "Only to retrieve results for specific subnational regions.")
dfs[8,]<-c("Estimation, simulation, and aggregation",
           "High density clusters",
           "HDENS_CLST_2021",
           "Eurostat GISCO",
           "1km grid",
           "2021",
           "unknown",
           "We use this as input to create alternative urban area specifications")
dfs[9,]<-c("Estimation, simulation, and aggregation",
           "Urban clusters",
           "URBAN_CLST_2021",
           "Eurostat GISCO",
           "1km grid",
           "2021",
           "unknown",
           "We use this as input to create alternative urban area specifications")
dfs[10,]<-c("Estimation, simulation, and aggregation",
           "Nordic daily maximum temperature data",
           "NGCD_TX_type1_version_26.03_2024MMDD.nc",
           "COPERNICUS",
           "1km grid",
           "2024",
           "yearly (I think)",
           "We use this as input to identify which urban areas should be included in which month")
#print
kable(dfs)
# clean up
rm(dfs)
```

# 3. Import and format data

## 3.1 LCM & urban areas

We start by identifying cities in Norway that are counted within category "city" in the DGURBA dataset.
```{r,message = FALSE, results = 'hide'}
gc()
# read Grunnkartet -------------------------------------------------------------
lcm<-terra::rast(file.path(d_fp,"Norway_files","Grunnkart","Grunnkart_OT.tif"))

# read city shapefiles ---------------------------------------------------------
# Readme file in INCA's data repository states DEGURBA should be used for 2026
# reporting
cityb<-terra::vect(file.path(d_fp,"version_2026_7",
                             "inca_input_local_climate_regulation","Urban_Areas",
                             "2021","DEGURBA",
                             "DGURBA_RG_01M_2021_4258.shp"))
# MM: I checked the data and it has five categories. 1,2 and 3 relate to  degree 
# of urbanisation (1 is cities, 2 is towns and semi-dense areas, 3 is rural). 
# Additionally there is category 9 which exists only in Cyprus, so irrelevant 
# for us. And then there is category 0 which is used in lakes and such. Also not 
# relevant for us. So we subset the data to Norway only.
cityb<-cityb[cityb$CNTR_CODE == "NO",]
# And then we subset to feature category 1 urban only.
citytownb<-cityb[cityb$DGURBA %in% c(1,2),]
cityb<-cityb[cityb$DGURBA == 1,]
# For further operations we reproject the map to LCM's CRS and we add a 150m
# buffer to the LAU polygons, as suggested by the handbook
# INCA introduces the same buffer. See line 280 here:
# https://code.europa.eu/inca/gis-plugins/gis-plugins/-/blob/main/src/inca/localclimate/__init__.py?ref_type=heads
cityb <- terra::project(cityb,lcm)
cityb_buf <- terra::buffer(cityb,150)
citytownb <- terra::project(citytownb,lcm)
citytownb_buf <- terra::buffer(citytownb,150)
# save
terra::writeVector(cityb,file.path(rfp,"no_city.shp"),overwrite=T)
terra::writeVector(cityb_buf,file.path(rfp,"no_citywbuffer.shp"),overwrite=T)
terra::writeVector(citytownb,file.path(rfp,"no_city_town.shp"),overwrite=T)
terra::writeVector(citytownb_buf,file.path(rfp,"no_city_town_wbuffer.shp"),overwrite=T)
# print the city names
# unique(cityb$LAU_NAME)
```

As a result we get `r length(unique(cityb$LAU_NAME))` LAU in Norway that qualify as city. If wanted to include towns in the computation as well, we would have `r length(unique(citytownb$LAU_NAME))` LAU. If we count Sandnes to Stavanger, and Bærum, Rælingen and Lørenskog to Oslo, we have four cities in total, namely Bergen, Oslo, Stavanger, and Trondheim. Let's have a look at the land cover map for each of the LAU to be considered. We translate the land cover map to the ecosystem typology required by Eurostat. Category 10 (marine inlets and transitional waters) is missing in the Norwegian LCM. This does not matter for the ecosystem service of local climate regulation as we only consider cooling from vegetation, according to the handbook.

```{r,message = FALSE, results = 'hide'}
gc()
# then the recategorisation of grunnkartet
lcm_idf<-read.csv(file.path(d_fp,"Norway_files","Grunnkart",
                            "Lookup_grunnkart.csv"))
lcm_idf <- rbind(lcm_idf,c(NA,9900,"unclassified","uklassifisert"))
lcm_idf$landcover_id <- as.numeric(lcm_idf$landcover_id)
# make the vector of ecosystem types
estv<-c("Settlements and other artificial areas",
        "Cropland",
        "Grassland",
        "Forest and woodlands",
        "Heathlands and shrub",
        "Sparsely vegetated ecosystems",
        "Inland wetlands",
        "Rivers and canals",
        "Lakes and reservoirs",
        "Marine inlets and transitional waters",
        "Coastal beaches, dunes and wetlands",
        "Marine ecosystems")
# and add it to our lcm_idf
lcm_idf['es_type']<-NA
lcm_idf$es_type[lcm_idf$landcover_id %in% c(100,111,112,121,130,131,132,133,134,135,141,142,152)]<-estv[1] # settlem
lcm_idf$es_type[lcm_idf$landcover_id %in% c(200)]<-estv[2] # cropl
lcm_idf$es_type[lcm_idf$landcover_id %in% c(310,320)]<-estv[3] # grassl
lcm_idf$es_type[lcm_idf$landcover_id %in% c(400,410,420,440)]<-estv[4] # forest
lcm_idf$es_type[lcm_idf$landcover_id %in% c(520)]<-estv[5] # bush
lcm_idf$es_type[lcm_idf$landcover_id %in% c(600,611,623,631)]<-estv[6] # sparse
lcm_idf$es_type[lcm_idf$landcover_id %in% c(720)]<-estv[7] # wetlands
lcm_idf$es_type[lcm_idf$landcover_id %in% c(810)]<-estv[8] # rivers
lcm_idf$es_type[lcm_idf$landcover_id %in% c(910)]<-estv[9] # lakes
# category 10 (marine inlets and transitional waters) doesn't exist
lcm_idf$es_type[lcm_idf$landcover_id %in% c(1100)]<-estv[11] # coastal
lcm_idf$es_type[lcm_idf$landcover_id %in% c(1200)]<-estv[12] # marine ecosystems
#lcm_idf$es_type[lcm_idf$landcover_id %in% c(9900)]<-estv[12] # marine ecosystems
# make a numeric identifier also
lcm_idf$es_id <- match(lcm_idf$es_type, unique(estv))
# then add to raster
lev <- levels(lcm)[[1]]
lev$es_id <- lcm_idf$es_id[
  match(lev$Value, lcm_idf$landcover_id)
]
lev$es_type <- lcm_idf$es_type[
  match(lev$Value, lcm_idf$landcover_id)
]
levels(lcm) <- lev
# so here we can set which category we want to plot. 
# Can always see which categories we have available by calling cats(lcm)
activeCat(lcm) <- "es_type"
rm(lev)
```

Now as an alternative specification, we create an urban area definition similar to what Statistics Sweden has implemented. This is based on Eurostat raster data that takes in population. All 1km cells that host population of at least 5000 people are included. The area is then buffered by 1km, and subsequently intersected with city LAUs. So we stick to the same cities, but define the boundaries according to population density rather than administrative shapes.
```{r, message=FALSE, results='hide',warning=FALSE}
# read in high-density cluster raster data
hdcr<-terra::rast(file.path(d_fp,"..","GISCO","high_density","HDENS_CLST_2021.tif"))
# read in urban cluster raster data
ucr<-terra::rast(file.path(d_fp,"..","GISCO","urban_cluster","URBAN_CLST_2021.tif"))
# force into our geometry
hdcr<-resample(hdcr,lcm)
ucr<-resample(ucr,lcm)
# merge and categorise as binary
ccr<-hdcr+ucr
ccr[ccr>0]<-1
ccr[ccr==0]<-NA
# buffer by 1km
ccr_buf<-terra::buffer(ccr,1000)
ccr_buf <- ifel(ccr_buf,1,NA)
# make a polygon from it
urb_pol <- as.polygons(ccr_buf, dissolve = TRUE, na.rm = TRUE)
# and then mask our city LAUs with this polygon
cityb_buf_m<-intersect(cityb_buf,urb_pol)
# and then the same for towns
citytownb_buf_m<-intersect(citytownb_buf,urb_pol)
# tidy up
rm(hdcr,ucr,ccr,ccr_buf,urb_pol)
```

Now let's make a plot of all cities, with a blue shade for areas that are included in the stricter urban approach.
```{r,message = FALSE, results = 'hide', fig.height = 9}

# Make a plot of all cities
# import street data for better orientation
roads_list<-list()
for(i in 1:nrow(cityb_buf)){
  roads_list[[i]]<-terra::vect(file.path(here::here(),"ancillary_data",
                                         paste0("roads_",cityb_buf$LAU_ID[i],".shp")))
}
names(roads_list)<-cityb_buf$LAU_NAME
# Let's plot the data
for(i in 1:nrow(cityb_buf)){
  plot(mask(crop(lcm, cityb_buf[i,]),cityb_buf[i,]),main=paste(cityb_buf$LAU_NAME[i],"LCM"), legend=T)
  plot(cityb_buf_m[i,],col="dodgerblue",alpha=0.5, add=T)
  plot(roads_list[[i]],add=T)
}

```

Now next we need to identify which of these cities have experienced maximum daily temperatures of at least 25 degrees celsius in the reference year of 2024. For this purpose we use gridded COPERNICUS data on daily maximum temperatures. We aggregate this data to montly maps that show for each pixel the number of days on which temperatures exceeded 25 degrees. We then use this to filter for each month which urban areas have experienced any days with temperatures exceeding 25 degrees.
```{r,message = FALSE, warning=FALSE, results='hide'}
gc()
# make a function
mask_heat_fun<-function(inputurbanshape,lcm,heatrast){
  # Step 1: rasterise the inputshape
  cityr<-terra::rasterize(inputurbanshape,lcm,field="LAU_ID")
  # Step 2: align heatraster with lcm and make binary
  heatrast<-resample(heatrast,lcm,method="near")
  heatrast[heatrast>0]<-1
  heatrast[heatrast==0]<-NA
  # Step 3: Mask the urban raster by what falls under the heat days category
  cityrm<-mask(cityr,heatrast)
  # Step 4: Identify city codes that fall into the category
  cities <- unique(na.omit(values(cityrm)))
  citycodes <- levels(cityr)[[1]]$LAU_ID[levels(cityr)[[1]]$ID %in% cities]
  # Step 5: filter city shape accordingly
  outputurbanshape <- inputurbanshape[inputurbanshape$LAU_ID %in% citycodes,]
  # Step 6: return result and tidy up
  rm(cityr,cityrm,cities,citycodes)
  return(outputurbanshape)
  gc()
}

# apply function
cityb_l<-list()
cityb_buf_l<-list()
cityb_buf_m_l<-list()
citytownb_l<-list()
mdr=c(6,7,8,9)
for(j in 1:length(mdr)){
  cityb_l[[j]]<-mask_heat_fun(inputurbanshape=cityb,
                    lcm=lcm,
                    heatrast=terra::rast(file.path(here::here(),"ancillary_data",paste0("heatdays_",j,".tif")))
                    )
  cityb_buf_l[[j]]<-mask_heat_fun(inputurbanshape=cityb_buf,
                    lcm=lcm,
                    heatrast=terra::rast(file.path(here::here(),"ancillary_data",paste0("heatdays_",j,".tif")))
                    )
  cityb_buf_m_l[[j]]<-mask_heat_fun(inputurbanshape=cityb_buf_m,
                    lcm=lcm,
                    heatrast=terra::rast(file.path(here::here(),"ancillary_data",paste0("heatdays_",j,".tif")))
                    )
  citytownb_l[[j]]<-mask_heat_fun(inputurbanshape=citytownb,
                    lcm=lcm,
                    heatrast=terra::rast(file.path(here::here(),"ancillary_data",paste0("heatdays_",j,".tif")))
                    )
  gc()
  }
# identify citytowns that never get included
cni<-citytownb$LAU_NAME[!citytownb$LAU_NAME %in% unique(c(citytownb_l[[1]]$LAU_NAME,citytownb_l[[2]]$LAU_NAME,citytownb_l[[3]]$LAU_NAME,citytownb_l[[4]]$LAU_NAME))]
```

Non-matching CRS doesn't matter. Our analysis shows the following. Month June 2024 includes `r length(cityb_l[[1]]$LAU_ID)` cities, and `r length(citytownb_l[[1]]$LAU_ID)` cities + towns that have experienced temperatures exceeding 25 degrees. Month July includes `r length(cityb_l[[2]]$LAU_ID)` cities and `r length(citytownb_l[[2]]$LAU_ID)` cities + towns. Month August includes `r length(cityb_l[[3]]$LAU_ID)` cities and `r length(citytownb_l[[3]]$LAU_ID)` cities + towns. Month September includes `r length(cityb_l[[4]]$LAU_ID)` cities and `r length(citytownb_l[[4]]$LAU_ID)` cities + towns. So these are the areas we compute our cooling for, in each of the cases. Towns that do not get included in the computation in any month, because they did not experience days with temperatures exceeding 25 degrees, are: `r if(length(cni)==0) {"None"} else {paste(cni,collapse=",")}`.

Now, continuing on from here, we only analyse cities, because we do not have LST data covering towns as well. The extension of this quantification to towns is left for future versions of the quantification.

## 3.2 LST

The Eurostat handbook suggests to use an inclusive approach. We therefore pick all available months for the LST data, which we read in next. Croping and reprojection is done outside of this script, because it is a computationally time-intensive process.
```{r,message = FALSE, results = 'hide'}
gc()
# read LST raster file ---------------------------------------------------------
# Here we have data available for June, July, August, September, and we pick
# as described above.
lstr_6_no<-terra::rast(file.path(here::here(),"ancillary_data",
                                 "lstr_6_no.tif"))
lstr_7_no<-terra::rast(file.path(here::here(),"ancillary_data",
                                 "lstr_7_no.tif"))
lstr_8_no<-terra::rast(file.path(here::here(),"ancillary_data",
                                 "lstr_8_no.tif"))
lstr_9_no<-terra::rast(file.path(here::here(),"ancillary_data",
                                 "lstr_9_no.tif"))

# mask to city LAUs
lstr_6_no <- mask(crop(lstr_6_no,cityb_buf),cityb_buf)
lstr_7_no <- mask(crop(lstr_7_no,cityb_buf),cityb_buf)
lstr_8_no <- mask(crop(lstr_8_no,cityb_buf),cityb_buf)
lstr_9_no <- mask(crop(lstr_9_no,cityb_buf),cityb_buf)

## MM comment: The unit of the data is Kelvin. Vito has applied the scaling factor
# that is mentioned here: 
# https://www.usgs.gov/landsat-missions/landsat-collection-2-surface-temperature
# You can trace the step in line 131 here
# https://code.europa.eu/inca/gis-plugins/gis-plugins/-/blob/main/src/inca/localclimate/preprocessing/preprocess_LST.js?ref_type=heads
# Before running regressions, VITO changes the unit to celsuis. See line 195 here:
# https://code.europa.eu/inca/gis-plugins/gis-plugins/-/blob/main/src/inca/localclimate/__init__.py?ref_type=heads
# so we do that here too.
lstr_6_no <- lstr_6_no - 273.15
lstr_7_no <- lstr_7_no - 273.15
lstr_8_no <- lstr_8_no - 273.15
lstr_9_no <- lstr_9_no - 273.15

# Move LST data all into one stack
lstr_stack <- c(lstr_6_no,lstr_7_no,lstr_8_no,lstr_9_no)
rm(lstr_6_no,lstr_7_no,lstr_8_no,lstr_9_no)
```

Then compute the median across these maps. A problem I have identified is that the LST data has large gaps, and for some of the cities this implies that there are hardly any month-overlapping LST data available. In INCA this problem is not explicitly sorted, so the median LST values can be representing a very small proportion of the city. We apply a gap-filling method here, which uses cross-sectional variation of LST across pixels within-months plus temporal variation of the across-pixel-median LST across months. Each pixel that has missing LST data for some month, gets allocated a constructed LST value, based on how it compares to other pixels in months where we have data for it, and how median LST in the city changes across months.
```{r,message = FALSE, results = 'hide'}
gc()
# we keep all months in all cities, to make sure that numbers across cities
# are comparable. INCA also doesn't drop months where LST is below 25 degrees.

# we loop across cities, and then mosaic back together
mlstl<-list()
malstl<-list()
# the city loop
for(i in 1:nrow(cityb_buf)){
  # Step 1: Extract the city
  lst6<-crop(lstr_stack[[1]],cityb_buf[i,],mask=T)
  lst7<-crop(lstr_stack[[2]],cityb_buf[i,],mask=T)
  lst8<-crop(lstr_stack[[3]],cityb_buf[i,],mask=T)
  lst9<-crop(lstr_stack[[4]],cityb_buf[i,],mask=T)
  
  # Step 2: Get median LST temp across available data in the city
  mu6m <- median(values(lst6),na.rm=T)
  mu7m <- median(values(lst7),na.rm=T)
  mu8m <- median(values(lst8),na.rm=T)
  mu9m <- median(values(lst9),na.rm=T)
  
  # Step 3: Compute difference from each pixel to the median in the month ("pixel anomaly")
  a6m <- lst6 - mu6m
  a7m <- lst7 - mu7m
  a8m <- lst8 - mu8m
  a9m <- lst9 - mu9m
  
  # Step 4: Compute for each pixel the median anomly across months. Note that this
  # is based on available pixels. So if a certain pixel has only data in one month,
  # the result will be the anomaly measured in that month.
  aavg <- median(c(a6m,a7m,a8m,a9m), na.rm=TRUE)
  
  # Step 5: Construct a made up LST on the basis of the anomalies (average cross-
  #sectional variation) and the median temperature of the month (accounts for
  # temporal variation) 
  r6 <- mu6m + aavg
  r7 <- mu7m + aavg
  r8 <- mu8m + aavg
  r9 <- mu9m + aavg
  
  # Step 6: Fill the data gaps in each month with the made up data.
  alst6 <- cover(lst6, r6)
  alst7 <- cover(lst7, r7)
  alst8 <- cover(lst8, r8)
  alst9 <- cover(lst9, r9)
  
  # Step 6: Create a median LST map based on the "gap-filled" monthly LST data
  # now let's create the median maps. And also one based on the non-modified data
  malstl[[i]]<-median(c(alst6,alst7,alst8,alst9))
  mlstl[[i]]<-median(c(lst6,lst7,lst8,lst9),na.rm=T)
  
  # Step 7: clean up
  rm(lst6,lst7,lst8,lst9,mu6m,mu7m,mu8m,mu9m,a6m,a7m,a8m,a9m,aavg,r6,r7,r8,r9,
     alst6,alst7,alst8,alst9)
}

malst<-do.call(terra::mosaic,c(malstl,fun=mean))
mlst<-do.call(terra::mosaic,c(mlstl,fun=mean))
# align extents
malst<-terra::extend(malst,terra::ext(lcm))
mlst<-terra::extend(mlst,terra::ext(lcm))
# make a stack and call it mlstr_stack
mlstr_stack<-c(malst,mlst)
names(mlstr_stack)<-c("SSB","INCA")
# clean up
rm(malstl,mlstl,malst,mlst)
```

Then plot the ready LST data, to show what the gap-filling process does. First the modified data:
```{r,message = FALSE, results = 'hide', fig.height = 9}
gc()
# Let's plot the data
for(i in 1:nrow(cityb_buf)){
plot(mask(crop(lcm, cityb_buf[i,]),cityb_buf[i,]),main=paste(cityb_buf$LAU_NAME[i],"median LST SSB version"), legend=F)
plot(mask(crop(mlstr_stack[[1]], cityb_buf[i,]),cityb_buf[i,]),add=T)
plot(cityb_buf_m[i,],alpha=0.3,col="dodgerblue",add=T)
plot(roads_list[[i]],add=T)
}
```

And then the difference between the modifed (SSB) and the non-modified (INCA) version. The largest data gaps prevail in Sandnes and Bergen. Also some gaps for Trondheim and Stavanger. This is where the modification has its largest effect. 
```{r,message = FALSE, results = 'hide', fig.height = 7}
gc()
# Let's plot the data
for(i in 1:nrow(cityb_buf)){
plot(mask(crop(lcm, cityb_buf[i,]),cityb_buf[i,]),main=paste(cityb_buf$LAU_NAME[i],"median LST modification difference"), legend=F)
plot(mask(crop((mlstr_stack[[1]]-mlstr_stack[[2]]), cityb_buf[i,]),cityb_buf[i,]),range = c(-4, 4),add=T)
plot(cityb_buf_m[i,],alpha=0.3,col="dodgerblue",add=T)
plot(roads_list[[i]],add=T)
}
```

The modified version of the LST data is based on the original LST data where all pixels are available, and it uses the constructed LST data where the original data has gaps. Areas that are blanked out in all of the months of the original LST data cannot be filled with our modification method, and therefore remain. As you can see, this is the case for a large area of Trondheim (the same cloud blocking the satellite image each month? I don't believe it....), and some areas of Sandnes and Bergen.

In cities where LST data gaps are more common in hotter months, our method will increase the median LST, on average, compared to what INCA does. In cities where LST data gaps are more common in colder months, our method will decrease the median LST, on avaerage, compared to what INCA does. In cities where there are hardly ever any gaps (like Lørenskog and Rælingen), the modification makes hardly any difference.

## 3.3 EVAP & TCD

Then we read in the other raster files needed. Tree cover density and then evaporation. For evaporation we have the choice between different datasets. One covering July to August, one covering June to August, and a third covering June to September. There is no guidance in the Eurostat handbook on which data to choose. To match the LST data selection, we choose to use June to September.
```{r, message = FALSE, results = 'hide'}
gc()
# rough crop
norcrop <- ext(
  3.5e6, 5.5e6, # xmin, xmax
  3.5e6, 5.5e6 # ymin, ymax
)

# read EVAP raster file --------------------------------------------------------
evapr<-terra::rast(file.path(d_fp,"version_2026_7",
                           "inca_input_local_climate_regulation",
                           "Vegetation_Evapotranspiration","2024",
                           "interc_transp_2024_Jun-Jul-Aug-Sept.tif"))
# I have found one funny thing here. The metadata in this raster file includes a
# scale factor of 0.01. R terra automatically applies this to the values of the 
# file, so it contains realistic numbers. But the python code in INCA does not
# apply the scale factor. Therefore, numbers in INCA for evapotranspiration are 
# 100 times larger than here. For now, I manually change this here, so we have
# results as close as possible.

# rough crop
evapr_no <- crop(evapr,norcrop)
# reproject
evapr_no <- terra::project(evapr_no,crs(lcm),res=res(evapr_no),method="bilinear")
# mask to city lau
evapr_no <- mask(crop(evapr_no,cityb_buf),cityb_buf)
# evapr_no <- 100*evapr_no # this step to immitate what INCA does (but it actually
# is not correct)
# When plotting the resulting map we see that it is a bit patchy, too. Possibly
# revisit later.


# read TCD raster file ---------------------------------------------------------
tcdr<-terra::rast(file.path(d_fp,"version_2026_7",
                           "inca_input_local_climate_regulation",
                           "Tree_Cover_Density","2023",
                           "TCD_100m_2023_compressed.tif"))
# rough crop
tcdr_no <- crop(tcdr,norcrop)
# reproject
tcdr_no <- terra::project(tcdr_no,evapr_no)
# mask to city lau
tcdr_no <- mask(crop(tcdr_no,cityb_buf),cityb_buf)
# MM I have checked the resulting data and it looks fine. No gaps.

# reproject lst raster data ----------------------------------------------------
mlstr_stack<-terra::project(mlstr_stack,evapr_no)
lstr_stack<-terra::project(lstr_stack,evapr_no)
# tidy up
rm(tcdr,evapr)
```

## 3.4 Mask stack

Before we continue on with estimating our models, we first create a spatial mask for each lst version that represents the valid pixels we can base the model on.
```{r, message = FALSE, results = 'hide'}
mask_inca_no<-terra::ifel(!is.na(evapr_no) & !is.na(tcdr_no) & !is.na(mlstr_stack[["INCA"]]),1,NA)
mask_ssb_no<-terra::ifel(!is.na(evapr_no) & !is.na(tcdr_no) & !is.na(mlstr_stack[["SSB"]]),1,NA)

mask_stack<-c(mask_ssb_no,mask_inca_no)
names(mask_stack)<-c("SSB","INCA")
# then we need lcm, make sure it is aligned with the rest of the raster data
lcm2 <- resample(lcm,mask_stack,method='near')
#clean up
rm(mask_inca_no,mask_ssb_no)
gc()
```

## 3.5 Air temp calibration data

Read in a single dataframe containing weather station data. This is the training data from the US.
```{r, message = FALSE, results = 'hide'}
gc()
# read in weather station data
wstd<-read.csv(file.path(d_fp,"version_2026_7",
                           "inca_input_local_climate_regulation",
                           "Weather_station_data",
                           "StationNeighborhood2016.csv"))
```

## 3.6 Latitude

A map of latitude, needed for simulations.
```{r, message = FALSE, results = 'hide'}
# Project raster to geographic coordinates
tmp_ll <- project(lstr_stack[[1]], "EPSG:4326")
# Create latitude raster
lat_ll <- init(tmp_ll, "y")
# Back to the original grid
lat_r <- project(lat_ll, lstr_stack[[1]])
rm(tmp_ll,lat_ll)
```

## 3.7 Diagnostics

Here we analyse for each version of the model the amount of pixels and proportions of pixels that are valid, used, seasonally biased, and modified.
```{r, message = FALSE, results = 'hide'}
# create table format
dtable<-as.data.frame(matrix(ncol=9,nrow=nrow(cityb)))
colnames(dtable)<-c("Urban_def","LST_version","City","pixels","valid_pixels","used_valid_pixels","prop_validp_used","prop_validp_seasonb","prop_validp_modif")
dtable$City<-cityb$LAU_NAME

# now make a loop
areaversions<-c("INCA","SSB")
lstversions<-c("INCA","SSB")

dl<-list()
for(j in 1:length(lstversions)){
  dl[[j]]<-list()
  for(k in 1:length(areaversions)){
    dl[[j]][[k]]<-dtable
    dl[[j]][[k]]$Urban_def <- areaversions[k]
    dl[[j]][[k]]$LST_version<-lstversions[j]
    for(i in 1:nrow(cityb)){
      # choose respective urban area definition
      if(areaversions[k]=="INCA"){
        dstack<-c(crop(mask_stack[[j]],cityb_buf[i,],mask=T),
                  crop(lcm2,cityb_buf[i,],mask=T),
                  crop(median(lstr_stack,na.rm=T),cityb_buf[i,],mask=T),
                  crop(median(lstr_stack),cityb_buf[i,],mask=T))
      } else if(areaversions[k]=="SSB"){
        dstack<-c(crop(mask_stack[[j]],cityb_buf_m[i,],mask=T),
                  crop(lcm2,cityb_buf_m[i,],mask=T),
                  crop(median(lstr_stack,na.rm=T),cityb_buf_m[i,],mask=T),
                  crop(median(lstr_stack),cityb_buf_m[i,],mask=T))
      }
      # make it a dataframe
      names(dstack)<-c("mask","lcm","lst_wb","lst_full")
      ddf<-as.data.frame(dstack)
      # compute diagnostics
      # number of pixels
      d1<-nrow(ddf)
      # number of valid pixels
      d2<-nrow(ddf %>% filter(!lcm %in% c("Lakes and reservoirs","Rivers and canals","Marine ecosystems"))) # remove rivers, lakes, sea
      # number of used valid pixels
      d3<-nrow(ddf %>% filter(mask==1 & !lcm %in% c("Lakes and reservoirs","Rivers and canals","Marine ecosystems")))
      # proportion of valid pixels used
      d4<-round(d3/d2,digits=2)
      # proportion of used valid pixels seasonally biased / modified
      d5<-round(nrow(ddf %>% filter(mask==1 & !lcm %in% c("Lakes and reservoirs","Rivers and canals","Marine ecosystems") & is.na(lst_full))) / d3,digits=2)
      # paste in the right place
      dl[[j]][[k]]$pixels[i]<-d1
      dl[[j]][[k]]$valid_pixels[i]<-d2
      dl[[j]][[k]]$used_valid_pixels[i]<-d3
      dl[[j]][[k]]$prop_validp_used[i]<-d4
      dl[[j]][[k]]$prop_validp_seasonb[i]<-if(lstversions[j]=="INCA"){d5}else if(lstversions[j]=="SSB"){0}
      dl[[j]][[k]]$prop_validp_modif[i]<-if(lstversions[j]=="INCA"){0}else if(lstversions[j]=="SSB"){d5}
      # clean up
      rm(d1,d2,d3,d4,d5,ddf,dstack)
    }
  }
  dl[[j]]<-do.call("rbind",dl[[j]])
}
diagnosticsdf<-do.call("rbind",dl)
rm(dl,dtable,lstr_stack)

diagnosticsdf %>% filter(Urban_def==LST_version)
```

Print the resulting diagnostics table. We only show the two versions we report, i.e. the one for Eurostat based on city-LAUs and unmodified LST data, and the one for the national publication, based on population density reduced city-LAUs and modified LST data. To explain the table:
- pixels is the total number of pixels in the defined urban area. Given our 100m raster resolution it is equal to hectare.
- valid_pixels is the number of pixels remaining when we subtract rivers, lakes, and the sea
- used_valid_pixels is the number of pixels remaining when we consider data gaps
- prop_validp_used is the proportion of used pixels over valid pixels
- prop_validp_seasonb is the proportion of used pixels that are seasonally biased because LST data for some months is missing for them
- prop_validp_modif is the proportion of used pixels that we modify in order to reduce seasonal bias ("gap-filling")

```{r}
# print result table
kable(diagnosticsdf %>% filter(Urban_def==LST_version))
```


Then we kable the table and only plot the two versions we use. 

# 4. Estimate models of LST

## 4.1 Overview

In this first step we use gridded data to estimate a linear model of local surface temperature (LST) predicted by tree cover density (TCD) and evaporation. We estimate this model for each city, so it requires that our gridded data is cropped to city boundaries plus a buffer of 150 metres (as specified in Eurostat's handbook). The model is: $$ LST_{j} = \alpha_{1,j} + \beta_{1} TCD + \gamma_{1} EVAP $$ For each city, we save the estimated coefficients of the linear model.

## 4.2 Run regressions

We loop across cities and across available months here. In each city and month we run a regression where grid cells are units of observation. Predicting LST with TCD and EVAP. We store the resulting coefficients for each city and month.
```{r, message = FALSE, results = 'hide'}
gc()
# lsts to consider
mdr<-c("SSB","INCA")

# our dataframe for results
reg1df<-as.data.frame(matrix(nrow=1,ncol=14))
colnames(reg1df)<-c("LST_version","city","pixels_in_shp","p_pixels_used","p_pixels_rm",
                    "p_pixels_rm_due_na_lcm","p_pixels_rm_due_na_data","avg_lst",
                    "avg_tcd","avg_evap","est_intercept","est_tcd","est_evap",
                    "R2")
# function
first_regression_function<-function(cityshp,lst_rfile,tcd_rfile,evap_rfile){
  # Step 1: Crop the data to the right shape, stack it, and turn into df
  estack <- c(
    crop(lst_rfile, cityshp, mask = TRUE),
    crop(tcd_rfile, cityshp, mask = TRUE),
    crop(evap_rfile, cityshp, mask = TRUE))
  names(estack)<-c("lst","tcd","evap")
  edf<-as.data.frame(estack)
  clean_edf <- edf %>% drop_na()
  rm(estack)
  # Step 2: Run regression (if there is any data)
  if(nrow(clean_edf)>0){
    elm <- lm(lst ~ tcd + evap, data = edf)
    } else {
    elm <- NULL
    }
  # Step 3: Save relevant info
  rv<-vector()
  # total pixel number in shape
  rv[1] <- mask(crop(init(lst_rfile, 1), cityshp), cityshp) %>% ncell()
  # proportion of pixels used
  if(nrow(clean_edf)>0){
    rv[2] <- nobs(elm) / rv[1]
    } else {
    rv[2] <- 0
    }
  # proportion of pixels removed
  rv[3] <- 1 - rv[2]
  # proportion of pixels removed because outside of lcm data
  rv[4] <- 1 - (nrow(edf) / rv[1])
  # proportion of pixels removed because of data gaps (lst,tcd,evap)
  if(nrow(clean_edf)>0){
    rv[5] <- (nrow(edf) - nobs(elm)) / rv[1]
    } else {
    rv[5] <- 1
    }
  # land surface temp
  rv[6] <- mean(clean_edf$lst)
  # average tree cover density
  rv[7] <- mean(clean_edf$tcd)
  # average evapotranspiration
  rv[8] <- mean(clean_edf$evap)
  # model results
  if(nrow(clean_edf)>0){
    # estimated intercept
    rv[9] <- elm$coefficients[1]
    # estimated coefficient on tre cover density
    rv[10] <- elm$coefficients[2]
    # estimated coefficient on evaporation
    rv[11] <- elm$coefficients[3]
    # model fit
    rv[12] <- summary(elm)$r.squared
    } else {
    rv[c(9,10,11,12)] <- NA
    }
  # clean up and return result
  rm(edf,clean_edf,elm)
  return(rv)
}

## then loop across LST versions and cities (eurostat version)
rl<-list()
for(j in 1:length(mdr)){
  rl[[j]]<-reg1df
  for(i in 1:nrow(cityb_buf)){
    rl[[j]][i,1]<-mdr[j]
    rl[[j]][i,2]<-as.data.frame(cityb_buf[i,])$LAU_NAME
    rl[[j]][i,3:14]<-first_regression_function(cityshp = cityb_buf[i,],
                                lst_rfile = mlstr_stack[[j]],
                                tcd_rfile = tcdr_no,
                                evap_rfile = evapr_no)
  }
}
firstdf<-do.call("rbind",rl)
# add city codes
firstdf <- firstdf %>% 
  mutate(city_code = cityb_buf$LAU_ID[match(city,cityb_buf$LAU_NAME)]) %>%
  relocate(city_code,.after=city)
#clean up
rm(rl)

## then loop across months and cities (strict urban version)
rl<-list()
for(j in 1:length(mdr)){
  rl[[j]]<-reg1df
  for(i in 1:nrow(cityb_buf_m)){
    rl[[j]][i,1]<-mdr[j]
    rl[[j]][i,2]<-as.data.frame(cityb_buf_m[i,])$LAU_NAME
    rl[[j]][i,3:14]<-first_regression_function(cityshp = cityb_buf_m[i,],
                                lst_rfile = mlstr_stack[[j]],
                                tcd_rfile = tcdr_no,
                                evap_rfile = evapr_no)
  }
}
firstdf_n<-do.call("rbind",rl)
# add city codes
firstdf_n <- firstdf_n %>% 
  mutate(city_code = cityb_buf_m$LAU_ID[match(city,cityb_buf_m$LAU_NAME)]) %>%
  relocate(city_code,.after=city)
#clean up
rm(reg1df,rl)
```

Print the table, with rounded values (and some less interesting rows and columns removed). Here the Eurostat version.
```{r}
# print result table
kable(firstdf %>% 
        select(-c(p_pixels_rm,p_pixels_rm_due_na_lcm,p_pixels_rm_due_na_data)) %>%
        filter(LST_version=="INCA") %>%
        mutate(across(where(is.numeric), ~ round(.x, 2))))
```

And here the version with stricter urban definition.
```{r}
# print result table
kable(firstdf_n %>% 
        select(-c(p_pixels_rm,p_pixels_rm_due_na_lcm,p_pixels_rm_due_na_data)) %>%
        filter(LST_version=="SSB") %>%
        mutate(across(where(is.numeric), ~ round(.x, 2))))
```

# 5. Estimate a model of air temp

## 5.1 Overview

In this second step we use weather station data to estimate a linear model of maximum air temperature predicted by local surface temperature and latitude. Here is the unit of observation is weather stations. We save the estimated coefficients.

## 5.2 Run regression

A single regression to estimate a linear model. Predicting maximum air temperature with local surface temperature and latitude. The model we estimate is: $$ T^{air} = \beta_{2} LST + \gamma_{2}LAT $$

```{r, message = FALSE, results = 'hide'}
gc()
# So here we now estimate the model and store the coefficients and adjusted r squared
# filter the data to only feature cases of positive temperature (has no effect because the data doesn't contain the opposite case)
wlm <- lm(TempMax ~ st + Ycoor, data = wstd %>% filter(TempMax > 0))
alpha2h <- wlm$coefficients['(Intercept)']
beta2h <- wlm$coefficients['st']
gamma2h <- wlm$coefficients['Ycoor']
wlm_adjr2<-summary(wlm)$adj.r.squared
# tidy up
rm(wstd)
```

Here are model outputs:
```{r}
summary(wlm)
rm(wlm)
```

# 6. Simulate cooling

## 6.1 Overview

Here we now apply the estimated coefficients from the previous two steps to generate simulated data. We do this for each city. INCA uses a different method for city cooling than for cooling by ecosystem, aggregated by region. In the first case it bases simulated air temperature with vegetation on observed LST data, and in the latter case it bases simulated air temperature with vegetation on simulated LST data. The rationale behind using different method for two different aggregation levels is unclear to me. For consistency reasons, we choose to apply the approach based on simulated LST data, not least because we expect that comparing modelled LST to modelled LST (as opposed to comparing observed LST to modelled LST) reduces potential bias.

We first need to construct predicted land surface temperature $\widehat{LST}_{j,i}$ for each grid cell $i$ and month $j$ with our calibrated model coefficients and observed inputs for the right hand side.

$$ \widehat{LST}_{j,i} = \widehat{\alpha}_{1,j} + \widehat{\beta}_{1,j}TCD_{j,i} + \widehat{\gamma}_{1,j}EVAP_{ji}$$

Then we compute for each grid cell $i$ and month $j$ the expected air temperature $\widehat{T}^{air_{simLSTgreen}}_{j,i}$ given simulated conditions.

$$ \widehat{T}^{air_{simLSTgreen}}_{j,i} = \widehat{\alpha}_{2} + \widehat{\beta}_{2} \widehat{LST}_{j,i} + \widehat{\gamma}_{2} LAT{j,i} $$.

Then we compute for each grid cell $i$ and month $j$ the expected air temperature in a no-vegetation scenario, i.e. where $\widehat{LST}^{zero}_{j,i} = \widehat{\alpha}_{1,j}$ because $TCD = EVAP = 0$ in this case. So

$$ \widehat{T}^{air_{simLSTgray}}_{j,i} = \widehat{\alpha}_{2} + \widehat{\beta}_{2} \widehat{LST}^{zero}_{j,i} + \widehat{\gamma}_{2} LAT_{j,i} $$.

Note that $\widehat{LST}^{zero}_{j,i}$ is constant for each grid cell in a city-month, because we only get one $\widehat{\alpha}_{1,j}$ for each city-month. The cooling effect is then retrieved for each grid cell $i$ in city $j$ as

$$ \widehat{cooling_{j,i}} = \widehat{T}^{air_{simLSTgray}}_{j,i} - \widehat{T}^{air_{simLSTgreen}}_{j,i}$$.

Again, as for the city statistics, we are effectively subtracting a constant no-vegetation counterfactual from the simulated varying air temperature with vegetation. After plugging in for estimated grid-cell temperature in both secenarios with vegetation and without vegetation, the previous equation simplifies to:

$$ \widehat{cooling}_{j,i} = -\widehat{\beta}_{2} * (\widehat{\beta}_{1,j}TCD_{j,i} + \widehat{\gamma}_{1,j}EVAP_{j,i})$$

As you can see, we do not actually need to use the local surface temperature data rasters when computing the cooling rasters based on fully simulated data. The cooling calculation simplifies to being based on the coefficients from both first and second regressions, as well as the tree cover raster and the evapotranspiration raster.

Just for the sake of completeness, let me show what INCA does differently for its city statistics. So here we first compute for each grid cell $i$ and month $j$ the expected air temperature $\widehat{T}^{air_{obsLSTgreen}}_{j,i}$ given observed conditions.

$$ \widehat{T}^{air_{obsLSTgreen}}_{j,i} = \widehat{\alpha}_{2} + \widehat{\beta}_{2}LST_{j,i} + \widehat{\gamma}_{2} LAT{j,i} $$.

Then we compute for each grid cell $i$ and month $j$ the expected air temperature in a no-vegetation scenario, same as above. The cooling effect is then retrieved for each grid cell $i$ in city $j$ as

$$ \widehat{cooling^{alt.method}_{j,i}} = \widehat{T}^{air_{simLSTgray}}_{j,i} - \widehat{T}^{air_{obsLSTgreen}}_{j,i}$$.

After plugging in for estimated grid-cell temperature in both secenarios with vegetation and without vegetation, the previous equation simplifies to:

$$ \widehat{cooling^{alt.method}_{j,i}} = -\widehat{\beta}_{2} * (LST_{j,i} + \widehat{\alpha}_{1,j})$$

As the formulation shows, cooling is a function of slightly different inputs than above, and this creates inconsistent results between the city statistics and the regional-by-ecosystem statistics. As said, we stick to the previous method.

## 6.2 Compute cooling rasters

### 6.2.1 Eurostat version

So here we compute cooling rasters based on the method described in the previous section. This we implement as closely as possible to INCA, but without inheriting INCA's bug using wrong coefficients. We also exclude water ecosystems from the computation as instructed by the handbook (INCA does not do this). As the Norwegian LCM does not include ecosystem "Marine inlets" in the first place, this means that we only need to exclude "Rivers and canals", "Lakes and reservoirs", and "Marine ecosystems".
```{r,message = FALSE, results = 'hide'}
gc()
### We run this as a loop across LST versions and cities. Following INCA's python implementation as close as possible
cooling_list_m <- vector("list", length(mdr))
for(j in 1:length(mdr)){
  # make a list for the city-wise cooling rasters
  cooling_list <- vector("list", nrow(cityb_buf))
  for(i in 1:nrow(cityb_buf)){
    # get lcm
    tmp_lcm <- crop(lcm2,cityb_buf[i,],mask=T)
    tmp_lcm[tmp_lcm %in% c(810,910,1200)]<-NA # rivers, lakes, sea, are set to NA
    # get coefficients
    alpha1h_ <- firstdf$est_intercept[firstdf$LST_version==mdr[j] & firstdf$city==cityb_buf$LAU_NAME[i]]
    beta1h_ <- firstdf$est_tcd[firstdf$LST_version==mdr[j] & firstdf$city==cityb_buf$LAU_NAME[i]]
    gamma1h_ <- firstdf$est_evap[firstdf$LST_version==mdr[j] & firstdf$city==cityb_buf$LAU_NAME[i]]
    # create simulated lst rasters
    lst_sim_green_ <- alpha1h_ + beta1h_*crop(tcdr_no,cityb_buf[i,],mask=T) + gamma1h_*crop(evapr_no,cityb_buf[i,],mask=T)
    lst_sim_gray_ <- alpha1h_ + 0*crop(tcdr_no,cityb_buf[i,],mask=T) + 0*crop(evapr_no,cityb_buf[i,],mask=T)
    # create simulated air temp rasters
    t_air_green_ <- alpha2h + beta2h*lst_sim_green_ + gamma2h*crop(lat_r,cityb_buf[i,],mask=T)
    t_air_gray_ <- alpha2h + beta2h*lst_sim_gray_ + gamma2h*crop(lat_r,cityb_buf[i,],mask=T)
    # cooling
    cooling_ <- t_air_gray_ - t_air_green_
    # unlike in INCA (but as requested in the hanbook) we set cooling of water bodies to NA
    cooling_[is.na(tmp_lcm)] <- NA
    # save in stack
    cooling_list[[i]]<-cooling_
    # tidy up
    rm(alpha1h_,beta1h_,gamma1h_,lst_sim_gray_,lst_sim_green_,t_air_gray_,t_air_green_,cooling_,tmp_lcm)
  }
  # in this function now we put the cities together and impose the mean cooling for the overlapping boundaries
  cooling_list_m[[j]] <- do.call(terra::mosaic,c(cooling_list,fun=mean))
  # make sure extents align (when some urban areas are masked because of lack of heat, then it can happen that extent is smaller than usual)
  cooling_list_m[[j]] <- terra::extend(cooling_list_m[[j]],terra::ext(lcm))
  # save
  writeRaster(cooling_list_m[[j]],file.path(rfp,paste0("cooling_",mdr[j],".tif")),overwrite=T)
}
cooling_stack<-terra::rast(cooling_list_m)
names(cooling_stack)<-mdr
cooling_stack_e <- cooling_stack
rm(cooling_list_m,cooling_stack)
```

Plot the cooling raster underlying the Eurostat version.
```{r,message = FALSE, results = 'hide', fig.height = 8}
for(i in 1:nrow(cityb_buf)){
   plot(crop(lcm,cityb[i,],mask=T),legend=F,main=paste(cityb_buf$LAU_NAME[i],"cooling","INCA_LST"))
  plot(crop(cooling_stack_e[["INCA"]], cityb_buf[i,],mask=T),add=T,range=c(-1,4),legend=T)
  plot(roads_list[[i]],add=T)
}
```

### 6.2.2 SSB version

```{r,message = FALSE, results = 'hide'}
gc()
### We run this as a loop across LST versions and cities. Following INCA's python implementation as close as possible
cooling_list_m <- vector("list", length(mdr))
for(j in 1:length(mdr)){
  # make a list for the city-wise cooling rasters
  cooling_list <- vector("list", nrow(cityb_buf_m))
  for(i in 1:nrow(cityb_buf_m)){
    # get lcm
    tmp_lcm <- crop(lcm2,cityb_buf_m[i,],mask=T)
    tmp_lcm[tmp_lcm %in% c(810,910,1200)]<-NA # rivers, lakes, sea, are set to NA
    # get coefficients
    alpha1h_ <- firstdf_n$est_intercept[firstdf_n$LST_version==mdr[j] & firstdf_n$city==cityb_buf_m$LAU_NAME[i]]
    beta1h_ <- firstdf_n$est_tcd[firstdf_n$LST_version==mdr[j] & firstdf_n$city==cityb_buf_m$LAU_NAME[i]]
    gamma1h_ <- firstdf_n$est_evap[firstdf_n$LST_version==mdr[j] & firstdf_n$city==cityb_buf_m$LAU_NAME[i]]
    # create simulated lst rasters
    lst_sim_green_ <- alpha1h_ + beta1h_*crop(tcdr_no,cityb_buf_m[i,],mask=T) + gamma1h_*crop(evapr_no,cityb_buf_m[i,],mask=T)
    lst_sim_gray_ <- alpha1h_ + 0*crop(tcdr_no,cityb_buf_m[i,],mask=T) + 0*crop(evapr_no,cityb_buf_m[i,],mask=T)
    # create simulated air temp rasters
    t_air_green_ <- alpha2h + beta2h*lst_sim_green_ + gamma2h*crop(lat_r,cityb_buf_m[i,],mask=T)
    t_air_gray_ <- alpha2h + beta2h*lst_sim_gray_ + gamma2h*crop(lat_r,cityb_buf_m[i,],mask=T)
    # cooling
    cooling_ <- t_air_gray_ - t_air_green_
    # unlike in INCA (but as requested in the hanbook) we set cooling of water bodies to NA
    cooling_[is.na(tmp_lcm)] <- NA
    # save in stack
    cooling_list[[i]]<-cooling_
    # tidy up
    rm(alpha1h_,beta1h_,gamma1h_,lst_sim_gray_,lst_sim_green_,t_air_gray_,t_air_green_,cooling_,tmp_lcm)
  }
  # in this function now we put the cities together and impose the mean cooling for the overlapping boundaries
  cooling_list_m[[j]] <- do.call(terra::mosaic,c(cooling_list,fun=mean))
  # make sure extents align (when some urban areas are masked because of lack of heat, then it can happen that extent is smaller than usual)
  cooling_list_m[[j]] <- terra::extend(cooling_list_m[[j]],terra::ext(lcm))
  # save
  writeRaster(cooling_list_m[[j]],file.path(rfp,paste0("cooling_n_",mdr[j],".tif")),overwrite=T)
}
cooling_stack<-terra::rast(cooling_list_m)
names(cooling_stack)<-mdr
cooling_stack_n <- cooling_stack
rm(cooling_list_m,cooling_stack)
```

Here are the cooling plots
```{r,message = FALSE, results = 'hide', fig.height = 8}
j=1
for(i in 1:nrow(cityb_buf_m)){
  plot(crop(lcm,cityb[i,],mask=T),legend=F,main=paste(cityb_buf$LAU_NAME[i],"cooling","LST version",mdr[j]))
  plot(crop(cooling_stack_n[["SSB"]], cityb_buf[i,],mask=T),range=c(-1,4),add=T,legend=T)
  plot(roads_list[[i]],add=T)
}
```

And here some more plots, just to show the difference to the case where we do not use modifed LST data.
```{r,message = FALSE, results = 'hide', fig.height = 8}
for(i in 1:nrow(cityb_buf_m)){
  plot(crop(lcm,cityb[i,],mask=T),legend=F,main=paste(cityb_buf$LAU_NAME[i],"cooling","LST version difference"))
  plot(crop((cooling_stack_n[[1]]-cooling_stack_n[[2]]), cityb_buf[i,],mask=T),range=c(-0.3,0.3),add=T,legend=T)
  plot(roads_list[[i]],add=T)
}
```

## 6.3 Eurostat deliverables

### 6.3.1 Overview

So here we report:
-  Supply tables, by ecosystem, by NUTS2 region, and total NUTS0
-  Contribution tables, from ecosystems, by NUTS2 region, and total NUTS0
-  Use tables

We use the version based on non-modified LST data (as INCA uses it) and based on non-modified city-LAU-based urban areas. 

### 6.3.2 Supply table

So here now we aggregate the cooling rasters by NUTS2 area, NUTS0 area, and ecosystem type. The 150m buffers imply that those NUTS2 areas that do not include city LAUs but that do have a border with city LAUs, do include cooling pixels. INCA does not address this problem. We set those cases to zero.

The indicators we report in our supply and use tables are what INCA calls "simple average" for the by-ecosystem values and what INCA calls "weighted average" for cooling across all ecosystem types.
```{r,message = FALSE, results = 'hide'}
gc()
# now cooling by nuts and ecosystem
# Import Norway NUTs2 shape
n2shp<-terra::vect(file.path(d_fp,"Norway_files","NUTS2021_NO_LVL2_wosvalbard","NUTS2021_NO_LVL2_wos.shp"))
n2shp <- project(n2shp,crs(lcm))
# rasterize the shape
n2r_no <- terra::rasterize(n2shp,mask_stack,field="NUTS_ID")
# now make a loop across lst versions
stl<-list()
for(j in 1:length(mdr)){
  # Step 1: Make a stack of the relevant data and move it into a dataframe
  astack <- c(
  terra::crop(cooling_stack_e[[j]],lcm2),
  n2r_no,
  lcm2)
  names(astack)<-c("cooling","region","lcm")
  adf<-as.data.frame(astack)
  clean_adf <- adf %>% drop_na()
  # Step 2: Retrieve all the different summary statistics we want
  # means by region and ecosystem type
  mean_by_region_es <- clean_adf %>% 
    group_by(region,lcm) %>% 
    summarise(cooling = mean(cooling))
  mean_by_es <- clean_adf %>% 
    select(-region) %>% 
    group_by(lcm) %>% 
    summarise(cooling = mean(cooling)) %>%
    mutate(region = "NO") %>%
    relocate(region,.before=lcm)
  # total means by NUTS2
  mean_by_region <- clean_adf %>% 
    group_by(region) %>% 
    summarise(cooling = mean(cooling)) %>% 
    mutate(lcm="All") %>%
    relocate(lcm,.after=region)
  # total mean for the whole country NUTS0
  mean_overall <- clean_adf %>% select(cooling) %>% 
    summarise(cooling=mean(cooling)) %>%
    mutate(region = "NO",lcm = "All") %>%
    relocate(cooling,.after =lcm)
  # put together
  mean_vals <- rbind(mean_by_es,mean_overall,mean_by_region_es,mean_by_region) %>% 
    filter(region !="NO09") %>% 
    pivot_wider(names_from=lcm,values_from = cooling) %>%
    left_join(n2shp %>% as.data.frame() %>% select(NUTS_ID,NUTS_NAME),
            by = join_by(region == NUTS_ID)) %>%
    rename(Name = NUTS_NAME) %>%
    relocate(Name,.after=region) %>%
    rename(Region = region)
  mean_vals$Name[mean_vals$Region=="NO"]<-"Norge"
  # put into our list
  stl[[j]]<-mean_vals
  # Step 3: clean up
  rm(astack,adf,clean_adf,mean_overall,mean_by_region,mean_by_es,mean_by_region_es,mean_vals)
}
names(stl)<-mdr
stl_e<-stl
rm(stl)
```

Here are the results.
```{r}
kable(stl_e[["INCA"]] %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```


### 6.3.3 Contribution table

To show how different ecosystems contribute to total cooling in each region, we construct seperate tables to show a decomposition of the (weighted) averages reported in the above section. This is what INCA calls "weighted average" for the by-ecosystem values. The decomposed values are not very insightful on their own, because they reflect the proportion of different ecosystems in each region (if the region consists predominantly of forest, forest will have the largest contribution to cooling). The cooling effectiveness of different ecosystems is better captured by the values reported in the previous section.
```{r,message = FALSE, results = 'hide'}
gc()
# decomposed total cooling by ecosystem for each region
# now make a loop across lst versions
cstl<-list()
for(j in 1:length(mdr)){
  # Step 1: Make a stack of the relevant data and move it into a dataframe
astack <- c(
  terra::crop(cooling_stack_e[[j]],lcm2),
  n2r_no,
  lcm2)
names(astack)<-c("cooling","region","lcm")
adf<-as.data.frame(astack)
clean_adf <- adf %>% drop_na()
# Step 2: Retrieve all the different summary statistics we want
# means by region and ecosystem type
mean_by_region_es <- clean_adf %>% 
  group_by(region,lcm) %>% 
  summarise(cooling = mean(cooling))
mean_by_es <- clean_adf %>% 
  select(-region) %>% 
  group_by(lcm) %>% 
  summarise(cooling = mean(cooling)) %>%
  mutate(region = "NO") %>%
  relocate(region,.before=lcm)
# weights
weights_by_region_es<-clean_adf %>% group_by(region) %>% count(lcm) %>% mutate(weight = n/sum(n)) %>% ungroup()
weights_overall_es<-clean_adf %>% count(lcm) %>% mutate(weight = n/sum(n)) %>% ungroup()

# put together
decomp_by_region_es <- mean_by_region_es %>% left_join(weights_by_region_es) %>%
  mutate(dcmp_cooling = cooling * weight)
decomp_overall_es <- mean_by_es %>% left_join(weights_overall_es) %>%
  mutate(dcmp_cooling = cooling * weight)
decomp_vals <- rbind(decomp_overall_es,decomp_by_region_es) %>% 
  filter(region !="NO09") %>% 
  select(-c(n,weight,cooling)) %>%
  pivot_wider(names_from=lcm,values_from = dcmp_cooling) %>%
  left_join(n2shp %>% as.data.frame() %>% select(NUTS_ID,NUTS_NAME),
            by = join_by(region == NUTS_ID)) %>%
  rename(Name = NUTS_NAME) %>%
  relocate(Name,.after=region) %>%
  rename(Region = region) %>%
  mutate(All = rowSums(across(`Settlements and other artificial areas`:`Coastal beaches, dunes and wetlands`)))
decomp_vals$Name[decomp_vals$Region=="NO"]<-"Norge"
  # put into our list
  cstl[[j]]<-decomp_vals
  # Step 3: clean up
  rm(astack,adf,clean_adf,mean_by_es,mean_by_region_es,decomp_vals,decomp_overall_es,decomp_by_region_es,weights_overall_es,weights_by_region_es)
}
names(cstl)<-mdr
cstl_e<-cstl
rm(cstl)
```

Here are the results
```{r}
kable(cstl_e[["INCA"]] %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```


### 6.3.4 Use table

```{r}
utl<-list()
for(j in 1:length(mdr)){
  utl[[j]]<-stl_e[[j]] %>% 
    select(c(Region,Name,All)) %>% 
    mutate(`Intermediate consumption by industries` = NA) %>%
    mutate(`Government final consumption` = NA) %>%
    mutate(`Households final consumption` = All) %>%
    mutate(`Gross capital formation` = NA) %>%
    mutate(Exports = NA) %>%
    rename(Total = All)
}
names(utl)<-mdr
utl_e<-utl
rm(utl)
```

Here are the results.
```{r}
kable(utl_e[["INCA"]] %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```


## 6.4 National deliverables

### 6.4.1 Overview

Here we report:
-  Supply tables, by ecosystem, and by city
-  Contribution tables, by ecosystem, and by city

Here we use the version based on modified LST data and modified urban areas.

### 6.4.2 Supply table

Same as in the previous section, but based on stricter urban definition, based on modified LST data, and aggregation to cities, rather than NUTS shapes.
```{r,message = FALSE, results = 'hide'}
gc()
# rasterize the city shape
cityr_no <- terra::rasterize(cityb_buf_m,mask_stack,field="LAU_ID")
# need to extend the stricter version so it is as large as the previous map
# this just adds NA colums on the right and on the left.
cooling_stack_n <- extend(
  cooling_stack_n,
  ext(cooling_stack_e)
)
# now make a loop across LST versions
stl<-list()
for(j in 1:length(mdr)){
  # Step 1: Make a stack of the relevant data and move it into a dataframe
  astack <- c(
  terra::crop(cooling_stack_n[[j]],lcm2),
  cityr_no,
  lcm2)
  names(astack)<-c("cooling","city","lcm")
  adf<-as.data.frame(astack)
  clean_adf <- adf %>% drop_na()
  # Step 2: Retrieve all the different summary statistics we want
  # means by city and ecosystem type
  mean_by_city_es <- clean_adf %>% 
    group_by(city,lcm) %>% 
    summarise(cooling = mean(cooling))
  # total means by NUTS2
  mean_by_city <- clean_adf %>% 
    group_by(city) %>% 
    summarise(cooling = mean(cooling)) %>% 
    mutate(lcm="All") %>%
    relocate(lcm,.after=city)
  # put together
  mean_vals <- rbind(mean_by_city_es,mean_by_city) %>% 
    pivot_wider(names_from=lcm,values_from = cooling) %>%
    left_join(cityb_buf_m %>% as.data.frame() %>% select(LAU_ID,LAU_NAME),
            by = join_by(city == LAU_ID)) %>%
    rename(Name = LAU_NAME) %>%
    relocate(Name,.after=city) %>%
    rename(City = city) %>%
    arrange(match(Name, cityb_buf_m$LAU_NAME))
  # put into our list
  stl[[j]]<-mean_vals
  # Step 3: clean up
  rm(astack,adf,clean_adf,mean_by_city,mean_by_city_es,mean_vals)
}
names(stl)<-mdr
stl_n<-stl
rm(stl)
```

Here are the results.
```{r}
kable(stl_n[["SSB"]] %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```

### 6.4.3 Contribution table

Same as in the previous version, but now based on strict urban areas.
```{r,message = FALSE, results = 'hide'}
gc()
# decomposed total cooling by ecosystem for each region
# now make a loop across months
cstl<-list()
for(j in 1:length(mdr)){
  # Step 1: Make a stack of the relevant data and move it into a dataframe
astack <- c(
  terra::crop(cooling_stack_n[[j]],lcm2),
  cityr_no,
  lcm2)
names(astack)<-c("cooling","city","lcm")
adf<-as.data.frame(astack)
clean_adf <- adf %>% drop_na()
# Step 2: Retrieve all the different summary statistics we want
# means by city and ecosystem type
mean_by_city_es <- clean_adf %>% 
  group_by(city,lcm) %>% 
  summarise(cooling = mean(cooling))
# weights
weights_by_city_es<-clean_adf %>% group_by(city) %>% count(lcm) %>% mutate(weight = n/sum(n)) %>% ungroup()

# put together
decomp_by_city_es <- mean_by_city_es %>% left_join(weights_by_city_es) %>%
  mutate(dcmp_cooling = cooling * weight)
decomp_vals <- rbind(decomp_by_city_es) %>% 
  select(-c(n,weight,cooling)) %>%
  pivot_wider(names_from=lcm,values_from = dcmp_cooling) %>%
  left_join(cityb_buf_m %>% as.data.frame() %>% select(LAU_ID,LAU_NAME),
            by = join_by(city == LAU_ID)) %>%
  rename(Name = LAU_NAME) %>%
  relocate(Name,.after=city) %>%
  rename(City = city) %>%
  mutate(All = rowSums(across(`Settlements and other artificial areas`:`Coastal beaches, dunes and wetlands`),na.rm=T)) %>%
    arrange(match(Name, cityb_buf_m$LAU_NAME))
  # put into our list
  cstl[[j]]<-decomp_vals
  # Step 3: clean up
  rm(astack,adf,clean_adf,mean_by_city_es,decomp_vals,decomp_by_city_es,weights_by_city_es)
}
names(cstl)<-mdr
cstl_n<-cstl
rm(cstl)
```

Here are the contribution results.
```{r}
kable(cstl_n[["SSB"]] %>% mutate(across(where(is.numeric), ~ round(.x, 2))))
```

# 7. Ideas for improvements

Several ideas, including:

- Use alternative LST data with better coverage. The current Lst data have been masked to city LAUs. If we had unmasked data, we could include all towns in the computation as well. This would increase the sample from 8 municipalities to up to 75 municipalities. It would provide more detailed and regionally varying information, and it would also create more representative results at the aggregate NUTS2 and NUTS0 level.
- Include more controls in the regression models. To change the cooling results, the controls would have to capture something that is currently (unintentionally) captured by vegetation. Most importantly NDVI, which captures more types of vegetation than only trees. DESTATIS in Germany has found out that this additional predictor makes the largest difference.
- Once available, consider using European weather station data to estimate the coefficients of the second model.
